### 2023-03-03 manuscript/manuscript.qmd
started on figure outline

---
title: "Fluoroproj: Title to be decided later"
author: "Stephen D. Shivers, Sophie Fournier, Jakob Stankoski, Jeffrey W. Hollister, Betty J. Kreakie"
abstract: "This is our abstract.  Just seeing how it looks."
format: docx
bibliography: manuscript.bib
---

## Introduction

Possible Journals: Lake and Reservoir Management, PLOS One, PLOS Water, Ecosphere, F1000Research,

The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria.

## Methods

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## Results

Here are our results. Fear us.

Figure ideas:

scatterplots: ext chl/phyco on y, chl/phyco from each insturment on x, colors/shapes for each waterbody/culture

boxplots (need to be same units): first boxplot is extracted, compare to all others.

## Discussion

Discuss amongst yourselves.

## References

### 2023-03-22 R/figures.R
added some text to manuscript, and updated scripts

@@ -9,7 +9,8 @@ fp_data_no_culture <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>
   filter(field_dups < 4) %>%
   group_by(date, waterbody,instrument,method,variable,units) %>%
   summarize(avg_value = mean(avg_value)) %>%
-  ungroup()
+  ungroup() %>%
+  filter(!grepl("cult", waterbody))
 
 fp_data_culture_only <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
   filter(instrument != "uri fluoroquik" & !is.na(method)) %>%
@@ -24,7 +25,7 @@ fp_data_culture_only <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) 
   
 
 # Plot #1 - Extracted Chl vs All others Chl scatterplot matrix
-chla_compare_plot <- ext_vs_all_plot(fp_data, "chl", c("fresh", "extracted"),
+chla_compare_plot <- ext_vs_all_plot(fp_data_culture_only, "chl", c("fresh", "extracted"),
                                      c("algaetorch (µg/L)", "phycoprobe (µg/L)",
                                        "fluoroquik (µg/L)", "cyanofluor (rfu)",
                                        "trilogy in vivo (rfu)"))

### 2023-03-22 R/functions.R

@@ -221,7 +221,7 @@ clean_field <- function(df){
 
 ext_vs_all_plot <- function(fpdata, var, meth, x_order = NULL){
   
-  
+  browser()
   
   if(var == "chl"){
     xvar <- sprintf("extracted chlorophyll (\u03BCg/L)")
@@ -255,7 +255,7 @@ ext_vs_all_plot <- function(fpdata, var, meth, x_order = NULL){
     facet_wrap(instrument_unit ~ ., scales = "free", strip.position = "left") +
     theme_ipsum_rc() +
     scale_color_brewer(type = "qual", palette = "Set1") +
-    scale_x_continuous(breaks = my_breaks) +
+    #scale_x_continuous(breaks = my_breaks) +
     theme(axis.title.y = element_blank(), 
           strip.text.y.left = element_text(angle=90, hjust = 1, size = 14), 
           strip.placement = "outside",

### 2023-03-31 manuscript/manuscript_sds.qmd
updated qmd manuscript

---
title: "Fluoroproj: Title to be decided later"
author: "Stephen D. Shivers, Sophie Fournier, Jakob Stankoski, Jeffrey W. Hollister, Betty J. Kreakie"
abstract: "This is our abstract.  Just seeing how it looks."
format: docx
bibliography: manuscript.bib
---

## Introduction

Possible Journals: Lake and Reservoir Management, PLOS One, PLOS Water, Ecosphere, F1000Research,

The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria.

## Methods

Equipment used for comparison

The following five fluorometers were used during this study: Turner Trilogy (in vivo chlorophyll module), Turner FluoroSense, Turner Cyanofluor, bbe AlgaeTorch, and bbe PhycoProbe. Measurements from these fluorometers were compared with the results of a solvent-based extraction for chlorophyll and phycocyanin using a Turner Trilogy with a chla-na and orange module, respectively.

Field Sampling Methods

Field samples were collected from eight ponds in Rhode Island during October and November 2021. Two liter surface samples were collected in triplicate (six total liters) by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment. If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles. Samples were collected in acid-washed 1 L amber bottles and placed in a cooler with ice. At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.

Analysis of field samples

Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction. Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in-vivo module. Chlorophyll a extraction using 90% acetone followed after a 20 minute period in a sonicating water bath. Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer. Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).

Analysis of lab grown culture

Two strains of cyanobacteria (*Microcystis* *aeruginosa* and *Anabaena* *flos-aquae)* and green algae (*Pediastrum boryanum* and *Pandorina morum*), purhcased from the University of Texas Culture Collection of Algae, have been maintained since August 2022.

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## Results

Here are our results. Fear us.

Figure ideas:

scatterplots: ext chl/phyco on y, chl/phyco from each insturment on x, colors/shapes for each waterbody/culture

boxplots (need to be same units): first boxplot is extracted, compare to all others.

## Discussion

Discuss amongst yourselves.

## References

### 2023-03-31 manuscript/manuscript_sds.qmd
updated manuscript

@@
 
-The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria.
+The goal of this paper is to compare different handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) in order to assess the feasibility of combining data from these various sources.
+
+Old Goal: The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria.
 

### 2023-03-31 manuscript/manuscript_sds.qmd
added outline to intro based on Betty

@@
 
-The goal of this paper is to compare different handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) in order to assess the feasibility of combining data from these various sources.
+HAB expansion is a problem and temporal nature means lots of groups have to be proactively involved in monitoring
+
+To respond to this problem industry has developed a variety of handhelds
+
+Taking individual fluorometer data across a region or between groups is difficult because these instruments all measure a little different
+
+The goal of this paper is to compare different handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) in order to provide a communications tool for groups using these different instruments.
 

### 2023-04-03 manuscript/manuscript.qmd
added some text to manuscript, and updated scripts
added notes and code to manusript for in situ/extract compare

@@
 
+This worked well until the fluorometer gave us problems.
+
+With culture, first look not very linear, but split out but green, cyan, and mixed each grouping shows linear relationship between extracted and in situ turner.
+Might not matter that different waterbodies would be in different green/cyano/mix classes.
+But the pattern between extracted and in situ/handheld flouros should be similar.
+Different pattern with any handheld/in situ would suggest it not perfomring well.
+
+```{r comparisons}
+library(readr)
+library(dplyr)
+library(ggplot2)
+
+prelim_ext_chl <- read_csv(here::here("data/raw/extracted chl/extracted_chl_fluororproj_prelimanary_test_2023_3_29.csv")) |>
+  group_by(waterbody) |>
+  summarize(extracted = mean(value, na.rm = TRUE)) |>
+  arrange(waterbody)
+prelim_in_viv <- read_csv(here::here("data/raw/trilogy in vivo/trilogy in vivo 2023_03_28_fresh.csv")) |>
+  mutate(waterbody = case_when(waterbody == "culture_aba" ~
+                                 "culture_ana",
+                               TRUE ~ waterbody))|>
+  group_by(waterbody) |>
+  summarize(in_vivo = mean(value, na.rm = TRUE)) |>
+  arrange(waterbody)
+
+all <- bind_cols (prelim_ext_chl, prelim_in_viv) |>
+  select(culture = waterbody...1, extracted, in_vivo) |>
+  filter(!is.na(culture)) |>
+  mutate(class = case_when(culture == "culture_ana" ~
+                             "cyano",
+                           culture == "culture_ana_pando" ~
+                             "mix",
+                           culture == "culture_mic" ~
+                             "cyano",
+                           culture == "culture_mic_ana" ~
+                             "cyano",
+                           culture == "culture_mic_pando" ~
+                             "mix",
+                           culture == "culture_pando" ~
+                             "greens",
+                           culture == "culture_pedia" ~
+                             "greens",
+                           culture == "culture_pedia_pando" ~
+                             "greens",
+                           culture == "culture_mic_ana_pando_pedia" ~
+                             "mix",
+                           TRUE ~ culture))
+
+ggplot(all, aes(x = extracted, y = in_vivo, color = class)) +
+  geom_point(size = 5)
+```
+
 ## Methods
@@
 1 + 1
+```
+
+```{r}
+library(dplyr)
+library(readr)
+read_csv("../data/cleaned_fluoroproj_data.csv")
 ```

### 2023-04-13 manuscript/manuscript_sds.qmd
updated introduction

@@
 
+Harmful algal blooms caused by cyanobacteria (cyanoHABs) are likely to increase in a warming world with greater inputs of nutrients by humans.
+These blooms may impact both human and environmental health, particularly if toxins are being produced.
+CyanoHAB blooms are often transient in nature, can be difficult to accurately quantify, and occur in diverse waterbodies across large spatial scales.
+Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions.
+As large, centrally managed monitoring efforts are expensive and time consuming to operate, a substantial portion of cyanoHAB monitoring occurs at the local level.
+As neighborhood and lake associations generally do not have the equipment or training to perform extractive analysis of cholorophyll a and/or phycocyanin (pigments found in all algae and specifically cyanobacteria, respectively), alternative measurement equipment has been developed.
+
 To respond to this problem industry has developed a variety of handhelds
 
+A variety of instruments have been developed to measure chlorophyll a and/or phycocyanin without an extraction step and have a wide range in cost.
+Most of these instruments rely on in vivo fluoroescense and can be used in the field or the lab, or a combination of both.
+
 Taking individual fluorometer data across a region or between groups is difficult because these instruments all measure a little different
+
+Field instruments may be placed directly in a waterbody or require water to be poured into a cuvette and then placed into a handheld fluorometer.
+Differences in measurement methods represent a potential source of error when comparing instruments.
+Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
 

### 2023-04-14 manuscript/manuscript_sds.qmd
updated introduction

@@
 
-A variety of instruments have been developed to measure chlorophyll a and/or phycocyanin without an extraction step and have a wide range in cost.
-Most of these instruments rely on in vivo fluoroescense and can be used in the field or the lab, or a combination of both.
+A variety of instruments have been developed to measure chlorophyll a and/or phycocyanin using in vivo fluorescence without an extraction step and have a wide range in cost.
+These instruments distinguish between the pigments based on their differing emission spectra.
+Some instruments (i.e.
+PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition.
+These instruments can be used in the field, the lab, or a combination of both.
 
@@
 Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
-
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light.
+Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior.
+Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
+Some instruments offer a yellow substance correction to help account for CDOM while others do not.
+How the different instruments handle these various potential problems can affect their accuracy, precision, and the ability to compare across different instrument types.
 The goal of this paper is to compare different handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) in order to provide a communications tool for groups using these different instruments.

### 2023-05-19 manuscript/manuscript_old_with_examp_ref.qmd
manuscript file cleanup

--- 2023-04-03 manuscript/manuscript.qmd
@@
 bibliography: manuscript.bib
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria.
+The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria [@Nojavan2019; @Hollister2016].
 
-This worked well until the fluorometer gave us problems.
+This worked well until the fluorometer gave us problems [@Hollister2016; @Walsh2017]
 
@@
 Might not matter that different waterbodies would be in different green/cyano/mix classes.
-But the pattern between extracted and in situ/handheld flouros should be similar.
+But the pattern between extracted and in situ/handheld flouros should be similar @Hodges2018.
 Different pattern with any handheld/in situ would suggest it not perfomring well.
@@
 
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
+When you click the **Render** button a document will be generated that includes both content and the output of embedded code . You can embed code like this:
 
@@
 
+![Instrument comparison for Chlorophyll](images/chla_instrument_compare.png){#fig-chla-comparison}
+
 ## Results
@@
 
-Discuss amongst yourselves.
+Discuss amongst yourselves. @fig-chla-comparison
 

### 2023-05-19 manuscript/manuscript.qmd
updates on manuscript to refelct field only.  chagnes to analysis.R

--- 2023-03-03 manuscript/manuscript.qmd
@@
 
-The goal of this paper is to compare accuracy and precision of a variety of fluorometers with respect to chlorophyll and phycocyanin across a variety of waterbodies as well as with lab grown cultures of green algae and cyanobacteria.
+<!-- HAB expansion is a problem and temporal nature means lots of groups have to be proactively involved in monitoring -->
+
+Harmful algal blooms caused by cyanobacteria (cyanoHABs) are likely to increase in a warming world with greater inputs of nutrients by humans.
+These blooms may impact both human and environmental health, particularly if toxins are being produced.
+CyanoHAB blooms are often transient in nature, can be difficult to accurately quantify, and occur in diverse waterbodies across large spatial scales.
+Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions.
+As large, centrally managed monitoring efforts are expensive and time consuming to operate, a substantial portion of cyanoHAB monitoring occurs at the local level.
+As neighborhood and lake associations generally do not have the equipment or training to perform extractive analysis of cholorophyll a and/or phycocyanin (pigments found in all algae and specifically cyanobacteria, respectively), alternative measurement equipment has been developed.
+
+<!--To respond to this problem industry has developed a variety of handhelds-->
+
+A variety of instruments have been developed to measure chlorophyll a and/or phycocyanin using in vivo fluorescence without an extraction step and have a wide range in cost.
+These instruments distinguish between the pigments based on their differing emission spectra.
+Some instruments (i.e.
+PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition.
+These instruments can be used in the field, the lab, or a combination of both.
+
+<!--Taking individual fluorometer data across a region or between groups is difficult because these instruments all measure a little different-->
+
+Field instruments may be placed directly in a waterbody or require water to be poured into a cuvette and then placed into a handheld fluorometer.
+Differences in measurement methods represent a potential source of error when comparing instruments.
+Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light.
+Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior.
+Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
+Some instruments offer a yellow substance correction to help account for CDOM while others do not.
+How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types.
+The goal of this paper is to compare different handhelds or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.
 
@@
 
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
+### Equipment used for comparison
 
-```{r}
-1 + 1
-```
+The following five fluorometers were used during this study: Turner Trilogy (in vivo chlorophyll module), Turner FluoroSense, Turner Cyanofluor, bbe AlgaeTorch, and bbe PhycoProbe.
+Measurements from these fluorometers were compared with the results of a solvent-based extraction for chlorophyll and phycocyanin using a Turner Trilogy with a chla-na and orange module, respectively.
 
-You can add options to executable code like this
+### Field Sampling Methods
 
-```{r}
-#| echo: false
-2 * 2
-```
+Field samples were collected from eight ponds in Rhode Island during October and November 2021.
+Two liter surface samples were collected in triplicate (six total liters) by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment.
+If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles.
+Samples were collected in acid-washed 1 L amber bottles and placed in a cooler with ice.
+At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.
 
-The `echo: false` option disables the printing of code (only output is displayed).
+### Analysis of field samples
+
+Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
+Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in-vivo module.
+Chlorophyll a extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
+Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer.
+Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).
 
@@
 
+Analysis ideas:
+
+-   Add in Phycotech data (all from 10-6)
+
+-   When did turnover happen? Might argue against using phycotech data for 10-21 and 11-09
+
+-   Maybe drop melville cuz to late/mismatch with phycotech.
+
 Figure ideas:
 
-scatterplots: ext chl/phyco on y, chl/phyco from each insturment on x, colors/shapes for each waterbody/culture
+-   scatterplots: ext chl/phyco on y, chl/phyco from each insturment on x, colors/shapes for each waterbody/culture
 
-boxplots (need to be same units): first boxplot is extracted, compare to all others.
+-   boxplots (need to be same units): first boxplot is extracted, compare to all others.
+
+-   Scatterplot of historic waterbody chlorophyll vs chlorophyll from each instrument
 

### 2023-05-19 R/analysis.R

@@ -1,8 +1,21 @@
 source(here::here("R/packages.R"))
 source(here::here("R/functions.R"))
 
-fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv"))
+fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
+  filter(!grepl("culture", waterbody),
+         !grepl("standard", waterbody),
+         waterbody != "windmist")
 
+
+
+
+
+
+
+
+
+
+# Beyond here be dragons
 inst_avg <- fp_data_wb %>%
   filter(method %in% c("frozen","extracted"), units == "µg/L", 
          variable == "chl") %>%

### 2023-06-02 manuscript/manuscript.qmd
More friday progress

@@
 
+### Phytoplankton Counts
+
+Add in stuff about sampling and Phycotech.
+
 ## Results
 
-Here are our results. Fear us.
+### Summary of Data
 
-Analysis ideas:
+Fig 1. Map of Field Sites
+
+### Chlorophyll
+
+Fig 2. Chlorophyll Scatterplot
+
+### Phycocyanin
+
+Fig 3. Phyco Scatterplot
+
+### Phycocyanin to Chlorophyll Ratio
+
+Fig 4. Ratio Scatterplot
+
+### Phytoplankton Counts
 
@@
 
-Figure ideas:
-
--   scatterplots: ext chl/phyco on y, chl/phyco from each insturment on x, colors/shapes for each waterbody/culture
-
--   boxplots (need to be same units): first boxplot is extracted, compare to all others.
-
--   Scatterplot of historic waterbody chlorophyll vs chlorophyll from each instrument
-
 ## Discussion

### 2023-06-02 R/analysis.R

@@ -8,10 +8,19 @@ fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
          waterbody != "windmist")
 
 # Figures
+# Have instruments in same spot on figure
+# color-blind friendly palette
+# Ratio Figure 
 chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
                                      c("algaetorch (µg/L)", "phycoprobe (µg/L)",
                                        "cyanofluor (rfu)","trilogy in vivo (rfu)"))
 
+phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted"),
+                                      c("algaetorch (µg/L of chlorophyll)", 
+                                        "phycoprobe (µg/L of chlorophyll)",
+                                        "fluoroquik (µg/L)", 
+                                        "fluorosense (µg/L)", 
+                                        "cyanofluor (rfu)"))
 
 
 

### 2023-06-30 R/analysis.R
working on grouped bar plot

@@ -8,10 +8,14 @@ fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
          waterbody != "windmist") %>%
   filter(!(waterbody %in% c("lower melville", "upper melville")))
 
+phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv"))
+
 # Figures
 # Have instruments in same spot on figure
 # color-blind friendly palette
 # Ratio Figure 
+# Grouped bar chart with waterbody on x, and bars for each division, relative and total biovolume - read from pre-sums division tab
+# Crazy idea, scatterplots from above but size of point from relative cyano biovolume
 chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
                                      c("algaetorch (µg/L)", "phycoprobe (µg/L)",
                                        "cyanofluor (rfu)",
@@ -25,8 +29,9 @@ phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted
 
 ratio_compare_plot <- ext_vs_all_plot(fp_data_wb, "pc:chl", c("fresh", "extracted"))
 
-# Grouped bar chart with waterbody on x, and bars for each division, relative and total biovolume - read from pre-sums division tab
-# Crazy idea, scatterplots from above but size of point from relative cyano biovolume
+division_bar_plot <- grouped_bar_plot(phycotech_data)
+
+
 
 
 

### 2023-06-30 R/data_prep.R

@@ -12,6 +12,8 @@ extracted_chla_data <- merge_extracted_chla() #On Feb 10, 2023 - blanks in new f
 extracted_phyco_data <- merge_extracted_phyco() #On Feb 10. 2023 - checking on solid standard, appeared to have typo (1044, should be 10044?)
 invivo_data <- merge_invivo()
 field_data <- read_csv(here("data/raw/field data.csv"))
+phycotech_data <- read_excel("data/phycotech/Pre_Sums_398_1_Algae_07012022.xlsx",
+                             sheet = "division_398_1_algae_07012022")
 
 # Clean up data
 handheld_data <- clean_handheld(handheld_data)
@@ -20,6 +22,7 @@ extracted_data <- bind_rows(extracted_chla_data, extracted_phyco_data)
 extracted_data <- clean_extracted(extracted_data)
 invivo_data <- clean_invivo(invivo_data)
 field_data <- clean_field(field_data)
+phycotech_data <- clean_phycotech(phycotech_data)
 
 
 fluoroproj_data <- bind_rows(handheld_data, phycoprobe_data, extracted_data,
@@ -66,3 +69,6 @@ fluoroproj_data_dups <- fluoroproj_data %>%
 
 write_csv(fluoroproj_data, here("data/cleaned_fluoroproj_data.csv"))
 write_csv(fluoroproj_data_dups, here("data/cleaned_fluoroproj_data_dups.csv"))
+
+
+# Phycotech data prep

### 2023-06-30 R/functions.R

@@ -421,3 +421,21 @@ fq_fresh_frozen_plot <- function(fpdata, var, x_order = NULL){
     labs(x = xvar, y = yvar)
   myplot
 }
+
+#' Clean Phycotech data
+clean_phycotech <- function(phycotech_df){
+  phycotech_df_clean <- phycotech_df %>%
+    select(waterbody = system_name, date = sample_date, division, 
+           concentration = concentration_natural_units_per_ml_,
+           relative_concentration, 
+           biovolume_concentration = total_biovolume_cubic_um_per_ml_,
+           relative_biovolume = relative_total_biovolume) %>%
+    mutate(date = ymd(date)) %>%
+    filter(date > "2021-09-01")
+  phycotech_df_clean
+}
+
+#' Make grouped bar plot
+grouped_bar_plot <- function(phycotech_df){
+  browser()
+}

### 2023-07-06 R/analysis.R
more updates

@@ -8,7 +8,8 @@ fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
          waterbody != "windmist") %>%
   filter(!(waterbody %in% c("lower melville", "upper melville")))
 
-phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv"))
+phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv")) %>%
+  filter(!(waterbody %in% c("Melville Pond")))
 
 # Figures
 # Have instruments in same spot on figure
@@ -16,6 +17,7 @@ phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv"))
 # Ratio Figure 
 # Grouped bar chart with waterbody on x, and bars for each division, relative and total biovolume - read from pre-sums division tab
 # Crazy idea, scatterplots from above but size of point from relative cyano biovolume
+# Compare phyco and chl to cell counts - maybe can make argument that cell counts bad, phyco good.
 chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
                                      c("algaetorch (µg/L)", "phycoprobe (µg/L)",
                                        "cyanofluor (rfu)",

### 2023-07-06 R/data_prep.R

@@ -69,6 +69,7 @@ fluoroproj_data_dups <- fluoroproj_data %>%
 
 write_csv(fluoroproj_data, here("data/cleaned_fluoroproj_data.csv"))
 write_csv(fluoroproj_data_dups, here("data/cleaned_fluoroproj_data_dups.csv"))
+write_csv(phycotech_data, "data/cleaned_phycotech_data.csv")
 
 
 # Phycotech data prep

### 2023-07-06 R/functions.R

@@ -430,12 +430,20 @@ clean_phycotech <- function(phycotech_df){
            relative_concentration, 
            biovolume_concentration = total_biovolume_cubic_um_per_ml_,
            relative_biovolume = relative_total_biovolume) %>%
-    mutate(date = ymd(date)) %>%
+    # TODO: Standardize names with our data
+    mutate(date = ymd(date),
+           waterbody = case_when(waterbody == "Yawagoo Pond" ~
+                                   "Yawgoo",
+                                 TRUE ~ waterbody)) %>%
     filter(date > "2021-09-01")
   phycotech_df_clean
 }
 
 #' Make grouped bar plot
 grouped_bar_plot <- function(phycotech_df){
-  browser()
+  #browser()
+  bar_plot <- phycotech_df %>%
+    ggplot(aes(x = waterbody, y = relative_biovolume, fill = division)) +
+    geom_bar(position = "dodge", stat = "identity")
+  bar_plot
 }

### 2023-07-07 manuscript/manuscript.qmd
working on grouped bar plot
making progress!

@@
 title: "Fluoroproj: Title to be decided later"
-author: "Stephen D. Shivers, Sophie Fournier, Jakob Stankoski, Jeffrey W. Hollister, Betty J. Kreakie"
+author: "Stephen D. Shivers, Jeffrey W. Hollister, Sophie Fournier, Jakob Stankoski, Betty J. Kreakie"
 abstract: "This is our abstract.  Just seeing how it looks."
@@
 
-Add in stuff about sampling and Phycotech.
+Add in stuff about sampling and Phycotech. Sampling dates of phytoplankton vs fluoro
 
@@
 
-Fig 2. Chlorophyll Scatterplot
+![Figure 2. Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg)
 
@@
 
-Fig 3. Phyco Scatterplot
-
-### Phycocyanin to Chlorophyll Ratio
-
-Fig 4. Ratio Scatterplot
+![Figure 3. Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg)
 
@@
 
--   Add in Phycotech data (all from 10-6)
+![Figure 4. Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg)
 
--   When did turnover happen? Might argue against using phycotech data for 10-21 and 11-09
+![Figure 5. Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg)
 
--   Maybe drop melville cuz to late/mismatch with phycotech.
+Table 1. Waterbody, Cyano Genus, Relative Biovolume Total Biovolume
+
+```{r read_tabl1, echo=FALSE, message=FALSE}
+table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) 
+```
+
+```{r show_table1, results='asis'}
+knitr::kable(table1)
+```
 
@@
 
+## Acknowledgements
+
+And special thanks to Gracie, Anastasia, and Lilac for emotional support during challenging analysis times.
+
 ## References

### 2023-07-07 R/analysis.R

@@ -9,7 +9,16 @@ fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
   filter(!(waterbody %in% c("lower melville", "upper melville")))
 
 phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv")) %>%
-  filter(!(waterbody %in% c("Melville Pond")))
+  filter(!(waterbody %in% c("Melville Pond"))) %>%
+  mutate(barplot_groups = case_when(division == "Euglenophyta" ~
+                                      "Other",
+                                    division == "Haptophyta" ~
+                                      "Other",
+                                    division == "Pyrrhophyta" ~
+                                      "Other", 
+                                    division == "Cryptophyta" ~
+                                      "Other",
+                                    TRUE ~ division))
 
 # Figures
 # Have instruments in same spot on figure
@@ -22,16 +31,26 @@ chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
                                      c("algaetorch (µg/L)", "phycoprobe (µg/L)",
                                        "cyanofluor (rfu)",
                                        "trilogy in vivo (rfu)"))
+ggsave(here::here("figures/fig2_chla_scatter.jpg"), chla_compare_plot, 
+       width = 7.5, height = 5.25, dpi = 300)
 
 phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted"),
                                       c("algaetorch (µg/L of chlorophyll)", 
                                         "phycoprobe (µg/L of chlorophyll)",
                                         "cyanofluor (rfu)", 
                                         "fluorosense (µg/L)"))
+ggsave(here::here("figures/fig3_phyco_scatter.jpg"), phyco_compare_plot, 
+       width = 7.5, height = 5.25, dpi = 300)
 
 ratio_compare_plot <- ext_vs_all_plot(fp_data_wb, "pc:chl", c("fresh", "extracted"))
 
-division_bar_plot <- grouped_bar_plot(phycotech_data)
+division_bar_plot_relative <- grouped_bar_plot(phycotech_data, "relative_biovolume")
+ggsave(here::here("figures/fig4_rel_bio_bar.jpg"), division_bar_plot_relative, 
+       width = 7.5, height = 5.25, dpi = 300)
+
[…]

### 2023-07-07 R/data_prep.R

@@ -14,6 +14,8 @@ invivo_data <- merge_invivo()
 field_data <- read_csv(here("data/raw/field data.csv"))
 phycotech_data <- read_excel("data/phycotech/Pre_Sums_398_1_Algae_07012022.xlsx",
                              sheet = "division_398_1_algae_07012022")
+phycotech_genus_data <- read_excel("data/phycotech/Pre_Sums_398_1_Algae_07012022.xlsx",
+                             sheet = "taxalevel_398_1_algae_07012022")
 
 # Clean up data
 handheld_data <- clean_handheld(handheld_data)
@@ -22,7 +24,9 @@ extracted_data <- bind_rows(extracted_chla_data, extracted_phyco_data)
 extracted_data <- clean_extracted(extracted_data)
 invivo_data <- clean_invivo(invivo_data)
 field_data <- clean_field(field_data)
+phycotech_cyano_data <- clean_phycotech_cyano(phycotech_genus_data)
 phycotech_data <- clean_phycotech(phycotech_data)
+
 
 
 fluoroproj_data <- bind_rows(handheld_data, phycoprobe_data, extracted_data,
@@ -70,6 +74,4 @@ fluoroproj_data_dups <- fluoroproj_data %>%
 write_csv(fluoroproj_data, here("data/cleaned_fluoroproj_data.csv"))
 write_csv(fluoroproj_data_dups, here("data/cleaned_fluoroproj_data_dups.csv"))
 write_csv(phycotech_data, "data/cleaned_phycotech_data.csv")
-
-
-# Phycotech data prep
+write_csv(phycotech_cyano_data, "data/cleaned_phycotech_cyano_data.csv")

### 2023-07-07 R/functions.R

@@ -439,11 +439,70 @@ clean_phycotech <- function(phycotech_df){
   phycotech_df_clean
 }
 
+#' Clean Phycotech data
+clean_phycotech_cyano <- function(phycotech_df){
+  
+  phycotech_df_clean <- phycotech_df %>%
+    select(waterbody = system_name, date = sample_date, division, family, genus, 
+           biovolume_concentration = total_biovolume_cubic_um_per_ml_,
+           relative_biovolume = relative_total_biovolume) %>%
+    # TODO: Standardize names with our data
+    mutate(date = ymd(date),
+           waterbody = case_when(waterbody == "Yawagoo Pond" ~
+                                   "Yawgoo",
+                                 TRUE ~ waterbody)) %>%
+    filter(date > "2021-09-01") %>%
+    filter(division == "Cyanophyta") %>%
+    group_by(waterbody,family, genus) %>%
+    summarize(biovolume = sum(biovolume_concentration, na.rm = TRUE),
+              relative_biovolume = sum(relative_biovolume, na.rm = TRUE)) %>%
+    ungroup() %>%
+    arrange(waterbody, desc(relative_biovolume)) %>%
+    mutate(genus = case_when(is.na(genus) ~ 
+                               "Unknwon Chroococcaceae",
+                             TRUE ~ genus))
+  phycotech_df_clean
+}
+
 #' Make grouped bar plot
-grouped_bar_plot <- function(phycotech_df){
-  #browser()
+grouped_bar_plot <- function(phycotech_df, yvar){
+  
+  yvar <- rlang::sym(yvar)
   bar_plot <- phycotech_df %>%
-    ggplot(aes(x = waterbody, y = relative_biovolume, fill = division)) +
-    geom_bar(position = "dodge", stat = "identity")
+    ggplot(aes(x = waterbody, y = !!yvar, fill = barplot_groups)) +
+    geom_bar(position = "dodge", stat = "identity") +
+    scale_fill_manual(values = c("brown","springgreen", "goldenrod", "cyan","grey50")) +
+    theme_ipsum_rc()
+    
   bar_plot
+}
+
+# Make scatterplot of fluoro concentrations vs counts
+flouro_vs_count_plot <- function(fluoro_df, phycotech_df){
+  browser()
+  fluoro_wb <- fluoro_df %>%
+    filter(variable %in% c("chl", "phyco"),
+           !(instrument == "trilogy" & units == "rfu")) %>%
[…]

### 2023-07-07 manuscript/manuscript.qmd
more local changes

@@
 ```{r read_tabl1, echo=FALSE, message=FALSE}
-table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) 
+table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) %>%
+  select(-family) %>%
+  mutate(biovolume = round(biovolume, 1),
+         relative_biovolume = round(relative_biovolume, 5))
 ```

### 2023-07-07 R/analysis.R

@@ -20,6 +20,8 @@ phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv")) %>%
                                       "Other",
                                     TRUE ~ division))
 
+
+  
 # Figures
 # Have instruments in same spot on figure
 # color-blind friendly palette
@@ -52,7 +54,7 @@ division_bar_plot_total <- grouped_bar_plot(phycotech_data, "biovolume_concentra
 ggsave(here::here("figures/fig5_total_bio_bar.jpg"), division_bar_plot_total, 
        width = 7.5, height = 5.25, dpi = 300)
 
-
+fluoro_count_scatter <- flouro_vs_count_plot(fp_data_wb, phycotech_data)
 
 
 

### 2023-07-10 R/analysis.R
First draft of paper figures (except map)

@@ -33,7 +33,7 @@ chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
                                      c("algaetorch (µg/L)", "phycoprobe (µg/L)",
                                        "cyanofluor (rfu)",
                                        "trilogy in vivo (rfu)"))
-ggsave(here::here("figures/fig2_chla_scatter.jpg"), chla_compare_plot, 
+ggsave(here::here("manuscript/images/fig2_chla_scatter.jpg"), chla_compare_plot, 
        width = 7.5, height = 5.25, dpi = 300)
 
 phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted"),
@@ -41,95 +41,24 @@ phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted
                                         "phycoprobe (µg/L of chlorophyll)",
                                         "cyanofluor (rfu)", 
                                         "fluorosense (µg/L)"))
-ggsave(here::here("figures/fig3_phyco_scatter.jpg"), phyco_compare_plot, 
+ggsave(here::here("manuscript/images/fig3_phyco_scatter.jpg"), phyco_compare_plot, 
        width = 7.5, height = 5.25, dpi = 300)
 
-ratio_compare_plot <- ext_vs_all_plot(fp_data_wb, "pc:chl", c("fresh", "extracted"))
+#ratio_compare_plot <- ext_vs_all_plot(fp_data_wb, "pc:chl", c("fresh", "extracted"))
 
 division_bar_plot_relative <- grouped_bar_plot(phycotech_data, "relative_biovolume")
-ggsave(here::here("figures/fig4_rel_bio_bar.jpg"), division_bar_plot_relative, 
+ggsave(here::here("manuscript/images/fig4_rel_bio_bar.jpg"), division_bar_plot_relative, 
        width = 7.5, height = 5.25, dpi = 300)
 
 division_bar_plot_total <- grouped_bar_plot(phycotech_data, "biovolume_concentration")
-ggsave(here::here("figures/fig5_total_bio_bar.jpg"), division_bar_plot_total, 
[…]

### 2023-07-10 R/functions.R

@@ -478,8 +478,9 @@ grouped_bar_plot <- function(phycotech_df, yvar){
 }
 
 # Make scatterplot of fluoro concentrations vs counts
-flouro_vs_count_plot <- function(fluoro_df, phycotech_df){
-  browser()
+flouro_vs_count_plot <- function(fluoro_df, phycotech_df, xvar = c("chlorophyll", "phycocyanin")){
+  
+  xvar <- match.arg(xvar)
   fluoro_wb <- fluoro_df %>%
     filter(variable %in% c("chl", "phyco"),
            !(instrument == "trilogy" & units == "rfu")) %>%
@@ -502,7 +503,32 @@ flouro_vs_count_plot <- function(fluoro_df, phycotech_df){
                                    "curran",
                                  TRUE ~ waterbody)) %>%
     mutate(waterbody = tolower(waterbody))
-  plot_df <- left_join(fluoro_wb, phyco_wb)
-  # Need to make plot same as extracted vs plots - have all instruments vs counts  
-  plot(plot_df$concentration, plot_df$phyco, pch = 19, col = "red", cex = 5)  
+  
+  if(xvar == "chlorophyll"){
+    plot_df <- left_join(fluoro_wb, phyco_wb) %>%
+      select(waterbody, instrument, yvar = chl, concentration) %>%
+      filter(!is.na(yvar))
+  } else if(xvar == "phycocyanin"){
+    plot_df <- left_join(fluoro_wb, phyco_wb) %>%
+      select(waterbody, instrument, yvar = phyco, concentration) %>%
+      filter(!is.na(yvar))
+  }
+  
+  myplot <- plot_df %>%
+    ggplot(aes(x = concentration, y = yvar, color = waterbody)) +
+    geom_point(size = 4) +
+    facet_wrap(instrument ~ ., scales = "free", strip.position = "left") +
+    theme_ipsum_rc() +
+    scale_color_viridis_d(option = "plasma") +
+    #scale_x_continuous(breaks = my_breaks) +
+    theme(axis.title.y = element_blank(), 
+          strip.text.y.left = element_text(angle=90, hjust = 1, size = 14), 
+          strip.placement = "outside",
+          axis.text.x = element_text(size = 14),
+          axis.title.x = element_text(size = 14, hjust = 0.5, vjust = -1),
+          legend.text=element_text(size = 14),
+          axis.text.y = element_text(size = 14),
+          legend.title = element_text(size = 14)) +
+    labs(x = "cyanobacterial cells/ml")
+  myplot
 }

### 2023-07-12 manuscript/manuscript.qmd
First draft of paper figures (except map)
added text to start results

@@
 
+Stats?  One-to-one lines? Regressions with r2?
+
+Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values.  The Algaetorch and Phycoprobe measurements were similar to and exhibited a strong relationship with extracted chlorophyll concentrations.  The Cyanofluor and Trilogy in vivo measurements increased as extracted chlorophyll concentrations increased but the relationship was not as strong compared to the other instruments, primarily resulting from differences in three of the six sites (Curran, Mashapaug, and Warwick).
+
+Phycocyanin measurements for the Fluorosense and the Cyanofluor had a similar relationship with extracted phycocyanin although the Fluorosense overestimated the concentrations.  The Algaetorch and Phycoprobe measurements were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll.
+
+Cyanophytes were the dominant taxa by relative biovolume at three sites (Curran, Mashapaug, and Warwick), Chrysophytes at two sites (Barber and Yawgoo), and Bacillariophytes at one site (Indian).  Biovolume concentrations were substantially higher at Mashapaug and Warwick.
+
 ### Summary of Data
@@
 
-![Figure 4. Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg)
+![Figure 4. Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){width="2250"}
 
 ![Figure 5. Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg)
+
+![Figure 6. Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg)
+
+![Figure 7. Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg)
 

### 2023-07-14 manuscript/manuscript.qmd
notes added to manuscript

@@
 
+<!-- Look at yellow substance and blank corrections on cyanfluor and in vivo (maybe others).  -->
+
 Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
@@
 
+### Statistical Analysis
+
+Make some comment about why no 1:1 line, using regression and r2 to compare sensors
+
 ## Results
 
-Stats?  One-to-one lines? Regressions with r2?
+Stats?  Regressions with r2?
 
@@
 ### Summary of Data
+
+<!-- TODO: Add Regression Lines to plots add R2-->
 
@@
 
-```{r read_tabl1, echo=FALSE, message=FALSE}
+```{r read_table1, echo=FALSE, message=FALSE}
 table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) %>%
@@
 
-Discuss amongst yourselves.
+Chlorophyll is not so good in some cases, but can be explained by phytoplankton community distributions
+
+Phycocyanin is good
+
+Don't read anything into absolute values (i.e. chlorophyll from one sensor not equal to chlorophyll from another sensor) across sensors, but general patterns are still very useful
+
+Phycocyanin very good proxy for cell counts.
+Suggest replacing cell counts for waterbody listing decisions with phycocyanin.
+Would likely need additional work to furhter explore relationship, build predictive models of cell coutns, etc.
 

### 2023-10-19 R/analysis.R
Figures almost good enough...  Good enough for review, methinks!

@@ -30,35 +30,38 @@ phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv")) %>%
 # Crazy idea, scatterplots from above but size of point from relative cyano biovolume
 # Compare phyco and chl to cell counts - maybe can make argument that cell counts bad, phyco good.
 chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
-                                     c("algaetorch chlorophyll (µg/L)", "phycoprobe chlorophyll (µg/L)",
-                                       "cyanofluor chlorophyll (rfu)",
-                                       "trilogy in vivo chlorophyll (rfu)"))
+                                     c("algaetorch chlorophyll\n(µg/L)", 
+                                       "phycoprobe chlorophyll\n(µg/L)",
+                                       "cyanofluor chlorophyll\n(rfu)",
+                                       "trilogy in vivo chlorophyll\n(rfu)"))
 ggsave(here::here("manuscript/images/fig2_chla_scatter.jpg"), chla_compare_plot, 
-       width = 7.5, height = 5.25, dpi = 300)
+       width = 8, height = 5.25, dpi = 300)
 
 phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted"),
-                                      c("algaetorch phycocyanin (µg/L of chlorophyll)", 
-                                        "phycoprobe phycocyanin (µg/L of chlorophyll)",
-                                        "cyanofluor phycocyanin (rfu)", 
-                                        "fluorosense phycocyanin (µg/L)"))
+                                      c("algaetorch phycocyanin\n(µg/L of chlorophyll)", 
+                                        "phycoprobe phycocyanin\n(µg/L of chlorophyll)",
+                                        "cyanofluor phycocyanin\n(rfu)", 
+                                        "fluorosense phycocyanin\n(µg/L)"))
 ggsave(here::here("manuscript/images/fig3_phyco_scatter.jpg"), phyco_compare_plot, 
-       width = 7.5, height = 5.25, dpi = 300)
+       width = 8, height = 5.25, dpi = 300)
 
 #ratio_compare_plot <- ext_vs_all_plot(fp_data_wb, "pc:chl", c("fresh", "extracted"))
 
[…]

### 2023-10-19 R/functions.R

@@ -262,7 +262,7 @@ ext_vs_all_plot <- function(fpdata, var, meth, x_order = NULL){
     filter(method %in% meth, instrument != "trilogy",
            variable == var) %>%
     left_join(extracted_data) %>%
-    mutate(instrument_unit = paste0(instrument, " ", varname, " (", units, ")")) 
+    mutate(instrument_unit = paste0(instrument, " ", varname, "\n(", units, ")")) 
   if(is.null(x_order)){
     x_order <- unique(plot_data$instrument_unit)
   }
@@ -290,7 +290,9 @@ ext_vs_all_plot <- function(fpdata, var, meth, x_order = NULL){
     group_by(instrument_unit) %>%
     mutate(r_square = summary(lm(avg_value ~ extracted_value))$r.squared,
            r_square = round(r_square, 2),
-           r_square_x = 5,
+           r_square_x = case_when(variable == "chl" ~ 10,
+                                  variable == "phyco" ~ 3,
+                                  TRUE ~ 10),
            r_square_y = ylim * 0.9) %>%
     ungroup() 
   
@@ -305,16 +307,14 @@ ext_vs_all_plot <- function(fpdata, var, meth, x_order = NULL){
     facet_wrap(instrument_unit ~ ., scales = "free", strip.position = "left") +
     theme_ipsum_rc() +
     scale_color_viridis_d(option = "plasma") +
-    #scale_y_continuous(limits = c(0, .data$ylim)) +
-    #scale_x_continuous(breaks = my_breaks) +
     theme(axis.title.y = element_blank(), 
-          strip.text.y.left = element_text(angle=90, hjust = 1, size = 14), 
+          strip.text.y.left = element_text(angle=90, hjust = 1, size = 11), 
           strip.placement = "outside",
-          axis.text.x = element_text(size = 14),
-          axis.title.x = element_text(size = 14, hjust = 0.5, vjust = -1),
-          legend.text=element_text(size = 14),
-          axis.text.y = element_text(size = 14),
-          legend.title = element_text(size = 14)) +
+          axis.text.x = element_text(size = 11),
+          axis.title.x = element_text(size = 11, hjust = 0.5, vjust = -1),
+          legend.text=element_text(size = 11),
+          axis.text.y = element_text(size = 11),
+          legend.title = element_text(size = 11)) +
     labs(x = xvar) + 
     geom_text(aes(x = r_square_x, y = r_square_y,
[…]

### 2023-10-19 manuscript/manuscript.qmd
results draft completed and methods for phytoplankton collection added

@@
 author: "Stephen D. Shivers, Jeffrey W. Hollister, Sophie Fournier, Jakob Stankoski, Betty J. Kreakie"
-abstract: "This is our abstract.  Just seeing how it looks."
+abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.
+A variety of instruments have been developed to measure cyanoHAB indicators (i.e.
+chlorophyll a or phycocyanin) that do not require an extraction process.
+We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll a and phycocyanin at six different waterbodies in Rhode Island.
+We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (r2 ranging from 0.78-0.96).
+We found less consistency between in-vivo measurements of chlorophyll a and the extracted results (r2 between 0.34 and 0.82).
+Some variability in the chlorophyll a results can be explained by differences in the phytoplankton community at the different sampling locations.
+Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.
+Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.
+Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.
+The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
 format: docx
@@
 
-Add in stuff about sampling and Phycotech. Sampling dates of phytoplankton vs fluoro
+Water samples for phytoplankton identification and enumeration were decanted from field collected samples into 250 ml amber HDPE bottles and were preserved with 25% glutaraldehyde.
+Samples were stored at 4°C and shipped the week after collection on ice to Phycotech, Inc.
+
+There was a temporal disconnect between some of the fluorometric analysis and the phytoplankton collection dates.
+All phytoplankton samples were collected on 2021-10-06.
+Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22; fluoromery was performed with 24 hours of sample collection.
 
@@
 
-Stats?  Regressions with r2?
+Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond).
+Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (Figure 2).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
 
-Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values.  The Algaetorch and Phycoprobe measurements were similar to and exhibited a strong relationship with extracted chlorophyll concentrations.  The Cyanofluor and Trilogy in vivo measurements increased as extracted chlorophyll concentrations increased but the relationship was not as strong compared to the other instruments, primarily resulting from differences in three of the six sites (Curran, Mashapaug, and Warwick).
+Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond).
+The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (Figure 3).
+Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin .
 
-Phycocyanin measurements for the Fluorosense and the Cyanofluor had a similar relationship with extracted phycocyanin although the Fluorosense overestimated the concentrations.  The Algaetorch and Phycoprobe measurements were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll.
+Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(Figure 4).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(Figure 5).
+Microcystis and Woronchinia were the dominant genera at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).
 
-Cyanophytes were the dominant taxa by relative biovolume at three sites (Curran, Mashapaug, and Warwick), Chrysophytes at two sites (Barber and Yawgoo), and Bacillariophytes at one site (Indian).  Biovolume concentrations were substantially higher at Mashapaug and Warwick.
+Extracted chlorophyll a was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59).
+The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (Figure 6).
+Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(Figure 7).
 
@@
 
-Chlorophyll is not so good in some cases, but can be explained by phytoplankton community distributions
+For some instruments, chlorophyll is good but for others it is not so good. Part of the variability can be explained by phytoplankton community distributions
 
-Phycocyanin is good
+Phycocyanin is good for all instruments.
 

### 2023-10-20 manuscript/manuscript.qmd
some jeff notes in discussion

@@
 
+Jeff's take home:
+
+1.) In general, in situ phycocyanin better match to extracted than in situ chlorophyll is to extracted. The cholorophylll relationship is driven by phytoplankton community make up
+
+2.) Extracted chlorophyll and phycocyanin, are great match to cell counts. Suggests pigments are a good proxy for cell counts and could be used in place of cell counts.
+
+3.) A discussion with water quality managers about using extracted or in situ phycocyanin as a replacement for cell counts might be warranted.
+
 For some instruments, chlorophyll is good but for others it is not so good. Part of the variability can be explained by phytoplankton community distributions

### 2023-10-20 manuscript/manuscript.qmd
manuscript updates

--- 2023-10-19 manuscript/manuscript.qmd
@@
 
-There was a temporal disconnect between some of the fluorometric analysis and the phytoplankton collection dates.
 All phytoplankton samples were collected on 2021-10-06.
@@
 
-Don't read anything into absolute values (i.e. chlorophyll from one sensor not equal to chlorophyll from another sensor) across sensors, but general patterns are still very useful
+Don't read anything into absolute values (i.e.
+chlorophyll from one sensor not equal to chlorophyll from another sensor) across sensors, but general patterns are still very useful.
+Need to know info about specific sites to be useful.
 
@@
 Would likely need additional work to furhter explore relationship, build predictive models of cell coutns, etc.
+
+Extracted chla and phyco good for cell counts. Outliers for instruments don't throw off either extracted measurements.
 

### 2023-10-23 R/analysis.R
figures ready for review.  Need to do study sites map...

@@ -9,7 +9,7 @@ fp_data_wb <- read_csv(here("data/cleaned_fluoroproj_data_dups.csv")) %>%
   filter(!(waterbody %in% c("lower melville", "upper melville")))
 
 phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv")) %>%
-  filter(!(waterbody %in% c("Melville Pond"))) %>%
+  filter(!(waterbody %in% c("Melville Pond", "Georgiaville Pond", "Slack Reservoir", "Stafford Pond"))) %>%
   mutate(barplot_groups = case_when(division == "Euglenophyta" ~
                                       "Other",
                                     division == "Haptophyta" ~
@@ -18,7 +18,14 @@ phycotech_data <- read_csv(here("data/cleaned_phycotech_data.csv")) %>%
                                       "Other", 
                                     division == "Cryptophyta" ~
                                       "Other",
-                                    TRUE ~ division))
+                                    TRUE ~ division)) %>%
+  mutate(waterbody = case_when(waterbody == "Barber Pond" ~ "barber",
+                               waterbody == "Indian Lake" ~ "indian",
+                               waterbody == "Yawagoo Pond" ~ "yawgoo",
+                               waterbody == "JL Curran Reservoir" ~ "curran",
+                               waterbody == "Mashapaug Pond" ~ "mashapaug",
+                               waterbody == "Warwick Pond" ~ "warwick",
+                               TRUE ~ waterbody))
 
 
   

### 2023-10-24 R/analysis.R
started on site map

@@ -42,7 +42,7 @@ chla_compare_plot <- ext_vs_all_plot(fp_data_wb, "chl", c("fresh", "extracted"),
                                        "cyanofluor chlorophyll\n(rfu)",
                                        "trilogy in vivo chlorophyll\n(rfu)"))
 ggsave(here::here("manuscript/images/fig2_chla_scatter.jpg"), chla_compare_plot, 
-       width = 8, height = 5.25, dpi = 300)
+       width = 8.5, height = 5.25, dpi = 300)
 
 phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted"),
                                       c("algaetorch phycocyanin\n(µg/L of chlorophyll)", 
@@ -50,25 +50,25 @@ phyco_compare_plot <- ext_vs_all_plot(fp_data_wb, "phyco", c("fresh", "extracted
                                         "cyanofluor phycocyanin\n(rfu)", 
                                         "fluorosense phycocyanin\n(µg/L)"))
 ggsave(here::here("manuscript/images/fig3_phyco_scatter.jpg"), phyco_compare_plot, 
-       width = 8, height = 5.25, dpi = 300)
+       width = 8.5, height = 5.25, dpi = 300)
 
 #ratio_compare_plot <- ext_vs_all_plot(fp_data_wb, "pc:chl", c("fresh", "extracted"))
 
 division_bar_plot_relative <- grouped_bar_plot(phycotech_data, "relative_biovolume")
 ggsave(here::here("manuscript/images/fig4_rel_bio_bar.jpg"), division_bar_plot_relative, 
-       width = 8, height = 5.25, dpi = 300)
+       width = 8.5, height = 5.25, dpi = 300)
 
 division_bar_plot_total <- grouped_bar_plot(phycotech_data, "biovolume_concentration")
 ggsave(here::here("manuscript/images/fig5_total_bio_bar.jpg"), division_bar_plot_total, 
-       width = 8, height = 5.25, dpi = 300)
+       width = 8.5, height = 5.25, dpi = 300)
 
 # As of 10/19/2023 still need to clean these up a bit.
 # consistency on y axis for each unit, x axis labels are scrunched
 fluoro_chl_count_scatter <- flouro_vs_count_plot(fp_data_wb, phycotech_data, "chlorophyll")
[…]

### 2023-10-24 R/functions.R

@@ -603,3 +603,19 @@ flouro_vs_count_plot <- function(fluoro_df, phycotech_df, xvar = c("chlorophyll"
                   label = paste0("R² = ", r_square)), color="black")
   myplot
 }
+
+map_field_sites <- function(){
+  ri <- st_read("data/ri.shp") |>
+    st_transform(4326)
+  ri_wb <- st_read("data/ri_lakes.shp") |>
+    st_transform(4326)
+  sites <- ri_wb |>
+    filter(GNIS_NAME %in% c("Barber Pond", "Curran Lower Reservoir", 
+                            "John L Curran Lower Reservoir", "Indian Lake",
+                            "Mashapaug Pond", "Warwick Pond", "Yawgoo Pond"))
+    
+  plot(st_geometry(ri))
+  plot(ri_wb, add = TRUE)
+  plot(sites, add = TRUE, col = "red")
+  
+}

### 2023-10-24 R/packages.R

@@ -11,3 +11,4 @@ library(cowplot)
 library(hrbrthemes)
 library(ggbeeswarm)
 library(readxl)
+library(sf)

### 2023-10-24 R/analysis.R
figures for review are done!

@@ -72,3 +72,7 @@ fluoro_phyco_count_scatter <- flouro_vs_count_plot(fp_data_wb, phycotech_data, "
 ggsave(here::here("manuscript/images/fig7_phyco_vs_cells.jpg"), fluoro_phyco_count_scatter, 
        width = 8.5, height = 5.25, dpi = 300)
 
+# Field Site Map
+field_sites <- map_field_sites() 
+ggsave(here::here("manuscript/images/fig1_map.jpg"), field_sites, 
+       width = 6, height = 8, dpi = 300)

### 2023-10-24 R/functions.R

@@ -605,17 +605,28 @@ flouro_vs_count_plot <- function(fluoro_df, phycotech_df, xvar = c("chlorophyll"
 }
 
 map_field_sites <- function(){
+  
   ri <- st_read("data/ri.shp") |>
     st_transform(4326)
   ri_wb <- st_read("data/ri_lakes.shp") |>
     st_transform(4326)
   sites <- ri_wb |>
-    filter(GNIS_NAME %in% c("Barber Pond", "Curran Lower Reservoir", 
+    filter(GNIS_NAME %in% c("Barber Pond", 
                             "John L Curran Lower Reservoir", "Indian Lake",
                             "Mashapaug Pond", "Warwick Pond", "Yawgoo Pond"))
     
-  plot(st_geometry(ri))
-  plot(ri_wb, add = TRUE)
-  plot(sites, add = TRUE, col = "red")
+  sites_gg <- ggplot(ri) +
+    geom_sf(size = 0.75, fill = "grey95", show.legend = FALSE) +
+    geom_sf(data = ri_wb, fill = "darkblue", color = NA, alpha = 0.5) +
+    geom_sf(data= st_centroid(sites), aes(color = GNIS_NAME), size = 2.5) +
+    scale_color_viridis_d(option = "plasma") +
+    scale_shape_manual(values = c(16,17)) +
+    scale_x_continuous(breaks = seq(-72.0, -71.0, by  = 0.2)) +
+    scale_y_continuous(breaks = seq(41.0, 42.0, by = 0.2)) +
+    theme_ipsum() +
+    theme(legend.position = "bottom", legend.title = element_blank())
+    #theme(legend.title = element_blank()) #+
+    #guides(color = guide_legend(nrow = 3))
+  sites_gg
   
 }

### 2023-10-24 R/packages.R

@@ -12,3 +12,4 @@ library(hrbrthemes)
 library(ggbeeswarm)
 library(readxl)
 library(sf)
+library(scales)

### 2023-11-17 manuscript/manuscript.qmd
started on site map
figures for review are done!
worked on discussion

@@
 
-Possible Journals: Lake and Reservoir Management, PLOS One, PLOS Water, Ecosphere, F1000Research,
+<!-- Possible Journals: Lake and Reservoir Management, PLOS One, PLOS Water, Ecosphere, F1000Research, -->
 
@@
 Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(Figure 4).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(Figure 5).
-Microcystis and Woronchinia were the dominant genera at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).
+Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).
 
@@
 
-Fig 1. Map of Field Sites
+![Fig 1. Map of Field Sites](images/fig1_map.jpg)
 
@@
 ```{r read_table1, echo=FALSE, message=FALSE}
+library(dplyr)
+library(readr)
 table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) %>%
@@
 ## Discussion
+
+The results from this study indicate that in-vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll.
+Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements.
+Instruments measuring phycocyanin were better able to handle differences in the phytoplankton community between these study ponds that cause chlorophyll measurements to be less strongly correlated with extracted results.
+Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.
+The relationship between phycocyanin and all fluorometers was strong.
+The Fluorosense had the strongest correlation even with the differences in phytoplankton concentrations and community dynamics between the waterbodies.
+However, the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L).
+Hambdhani et al.
+also found that the Fluorosense performed well but overestimated chlorophyll a at higher algal concentrations (Hambdhani et al.
+2021).
+Therefore, using this particular instrument to enforce set criteria across a variety of waterbodies would probably not be recommended without further research.
+The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations but the output from these instruments is not a direct measure of phycocyanin; instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria.
+The Cyanofluor output was in rfus and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study.
+A study by Thomson-Laing et al.
+did convert rfus to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (Thomson-Laing et al.
+2020).
+They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.
+All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll a.
+All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship.
+Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran.
+Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in-vivo concentrations of chlorophyll (Zamyadi et al .
+2012).
+Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments.
+A study by Silva et al.
+found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll a concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al.
+2016).
+This study supports the use of extracted chlorophll and phycocyanin as an alternative to cell counts for the study waterbodies.
+Currently, in vivo chlorophyll a is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments.
+Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts.
+There are a variety of instruments capable of measuring phycocyanin using in-vivo fluorometry across a spectrum of prices.
+We hope this study will encourage further research and consideration of using in-vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.
 

### 2023-11-29 manuscript/manuscript.qmd
Updates: Format, title, affiliations (i.e. all I could handle late in the day!)

@@
 ---
-title: "Fluoroproj: Title to be decided later"
-author: "Stephen D. Shivers, Jeffrey W. Hollister, Sophie Fournier, Jakob Stankoski, Betty J. Kreakie"
+title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
+author: 
+  - name: Stephen D. Shivers
+    orcid: 0000-0003-0096-7433
+    email: shivers.stephen@epa.gov
+    corresponding: true
+    affiliation:
+      - id: epa
+        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
+        city: Narragansett
+        state: RI
+        zip: 02882
+  - name: Jeffrey W. Hollister
+    orcid: 0000-0002-9254-9740
+    affiliation:
+      - ref: epa
+  - name: Sophie Fournier
+    affiliation:
+      - id: orise
+        name: Oak Ridge Institute for Science and Education
+        city: Narragansett
+        state: RI
+        zip: 02882
+  - name: Jakob Stankoski
+    affiliation:
+      - id: orau
+        name: Oak Ridge Associated Universities
+        city: Narragansett
+        state: RI
+        zip: 02882
+  - name: Betty J. Kreakie
+    orcid: 0000-0002-4982-7610
+    affiliation:
+      - ref: epa
 abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.
@@
 The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
-format: docx
+format: 
+    docx:
+      reference-doc: manuscript_template.docx
 bibliography: manuscript.bib
+#filters: 
+#  - authors-block
 ---
 
-## Introduction
+```{r, include = FALSE}
+knitr::opts_chunk$set(
+  eval = TRUE,
+  cache = TRUE,
+  tidy = FALSE,
+  message = FALSE,
+  fig.height = 4.8,
+  fig.width = 6,
+  dpi = 300
+)
+```
+
+# Introduction
 
@@
 
-## Methods
+# Methods
 
-### Equipment used for comparison
+## Equipment used for comparison
 
@@
 
-### Field Sampling Methods
+## Field Sampling Methods
 
@@
 
-### Analysis of field samples
+## Analysis of field samples
 
@@
 
-### Phytoplankton Counts
+## Phytoplankton Counts
 
@@
 
-### Statistical Analysis
+## Statistical Analysis
 
@@
 
-## Results
+# Results
 
@@
 
-### Summary of Data
+## Summary of Data
 
@@
 
-### Chlorophyll
+## Chlorophyll
 
@@
 
-### Phycocyanin
+## Phycocyanin
 
@@
 
-## Discussion
+# Discussion
 

### 2023-11-30 manuscript/manuscript.qmd
More formatting

@@
 A variety of instruments have been developed to measure cyanoHAB indicators (i.e.
-chlorophyll a or phycocyanin) that do not require an extraction process.
-We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll a and phycocyanin at six different waterbodies in Rhode Island.
+chlorophyll *a* or phycocyanin) that do not require an extraction process.
+We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.
 We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (r2 ranging from 0.78-0.96).
-We found less consistency between in-vivo measurements of chlorophyll a and the extracted results (r2 between 0.34 and 0.82).
-Some variability in the chlorophyll a results can be explained by differences in the phytoplankton community at the different sampling locations.
+We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (r2 between 0.34 and 0.82).
+Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.
 Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.
@@
 
-A variety of instruments have been developed to measure chlorophyll a and/or phycocyanin using in vivo fluorescence without an extraction step and have a wide range in cost.
+A variety of instruments have been developed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step and have a wide range in cost.
 These instruments distinguish between the pigments based on their differing emission spectra.
@@
 Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in-vivo module.
-Chlorophyll a extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
+chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
 Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer.
@@
 
-Make some comment about why no 1:1 line, using regression and r2 to compare sensors
+<!-- Jeff -->
+
+Make some comment about why no 1:1 line, using regression and r2 to compare sensors.
 
@@
 
-Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond).
-Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (Figure 2).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
-
-Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond).
-The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (Figure 3).
-Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin .
-
-Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(Figure 4).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(Figure 5).
-Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).
-
-Extracted chlorophyll a was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59).
-The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (Figure 6).
-Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(Figure 7).
-
 ## Summary of Data
 
-<!-- TODO: Add Regression Lines to plots add R2-->
+<!-- Jeff -->
+
+Need summary of sites and data
 
@@
 
+Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond).
+Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (Figure 2).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
+
 ![Figure 2. Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg)
@@
 ## Phycocyanin
+
+Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond).
+The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (Figure 3).
+Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin.
 
@@
 
+Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(Figure 4).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(Figure 5).
+Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).
+
 ![Figure 4. Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){width="2250"}
@@
 ![Figure 5. Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg)
+
+Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59).
+The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (Figure 6).
+Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(Figure 7).
 
@@
 Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.
+
 The relationship between phycocyanin and all fluorometers was strong.
@@
 Hambdhani et al.
-also found that the Fluorosense performed well but overestimated chlorophyll a at higher algal concentrations (Hambdhani et al.
+also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations (Hambdhani et al.
 2021).
 Therefore, using this particular instrument to enforce set criteria across a variety of waterbodies would probably not be recommended without further research.
+
 The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations but the output from these instruments is not a direct measure of phycocyanin; instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria.
@@
 They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.
+
 All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll a.
@@
 A study by Silva et al.
-found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll a concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al.
+found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al.
 2016).
+
 This study supports the use of extracted chlorophll and phycocyanin as an alternative to cell counts for the study waterbodies.
-Currently, in vivo chlorophyll a is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments.
+Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments.
 Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts.
@@
 
-Jeff's take home:
+# Summary
 
@@
 3.) A discussion with water quality managers about using extracted or in situ phycocyanin as a replacement for cell counts might be warranted.
-
-For some instruments, chlorophyll is good but for others it is not so good. Part of the variability can be explained by phytoplankton community distributions
-
-Phycocyanin is good for all instruments.
 
@@
 
-Extracted chla and phyco good for cell counts. Outliers for instruments don't throw off either extracted measurements.
-
 ## Acknowledgements

### 2023-11-30 manuscript/manuscript.qmd
updates
added in fig and table captions, more org.

@@
   message = FALSE,
+  warning = FALSE,
   fig.height = 4.8,
@@
 ```
+
+<!-- ADD IN CODE SO IT RE-RUNS ON RENDER -->
 
@@
 
-Field samples were collected from eight ponds in Rhode Island during October and November 2021.
+Field samples were collected from eight ponds in Rhode Island during October and November 2021 (@fig-map).
 Two liter surface samples were collected in triplicate (six total liters) by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment.
@@
 At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.
+
+![Field sampling locations](images/fig1_map.jpg){#fig-map width=100%}
 
@@
 
-## Statistical Analysis
+## Data Analysis
 
-<!-- Jeff -->
-
-Make some comment about why no 1:1 line, using regression and r2 to compare sensors.
+To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots to assess the fit and compare the different sensors.
+We derived the coefficient of determination from the regressions as a measure of fit and qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts.
+We do not compare slopes as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.
 
@@
 
-Need summary of sites and data
-
-![Fig 1. Map of Field Sites](images/fig1_map.jpg)
+Need summary of sites and dataFig 1. Map of Field Sites
 
@@
 Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond).
-Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (Figure 2).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
+Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
 
-![Figure 2. Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg)
+![Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg){#fig-chlascatter width=80%}
 
@@
 Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond).
-The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (Figure 3).
+The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter).
 Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin.
 
-![Figure 3. Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg)
+![Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width=80%}
 
@@
 
-Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(Figure 4).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(Figure 5).
-Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (Table 1).
+Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-relbv).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(@fig-bv).
+Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).
 
-![Figure 4. Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){width="2250"}
+![Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){#fig-relbv width=80%}
 
-![Figure 5. Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg)
+![Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg){#fig-bv width=80%}
 
 Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59).
-The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (Figure 6).
-Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(Figure 7).
+The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount).
+Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(@fig-phycocount).
 
-![Figure 6. Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg)
+![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg){#fig-chlcount width=80%}
 
-![Figure 7. Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg)
-
-Table 1. Waterbody, Cyano Genus, Relative Biovolume Total Biovolume
-
-```{r read_table1, echo=FALSE, message=FALSE}
-library(dplyr)
-library(readr)
-table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) %>%
-  select(-family) %>%
-  mutate(biovolume = round(biovolume, 1),
-         relative_biovolume = round(relative_biovolume, 5))
-```
-
-```{r show_table1, results='asis'}
-knitr::kable(table1)
-```
+![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg){#fig-phycocount width=80%}
 
@@
 
-## Acknowledgements
+# Acknowledgements
 
@@
 
-## References
+# Supplement
+
+```{r read_table1, echo=FALSE, message=FALSE}
+library(dplyr)
+library(readr)
+table1 <- read_csv(here::here("data/cleaned_phycotech_cyano_data.csv")) %>%
+  select(-family) %>%
+  mutate(biovolume = round(biovolume, 1),
+         relative_biovolume = round(relative_biovolume, 5))
+```
+
+```{r show_table1, results='asis', echo=FALSE}
+#| label: tbl-supp
+#| tbl-cap: "Waterbody, Cyano Genus, Relative Biovolume Total Biovolume"
+
+knitr::kable(table1)
+```
+
+# References

### 2023-11-30 manuscript/manuscript.qmd
fixed figure sizes

@@
 
-<!-- ADD IN CODE SO IT RE-RUNS ON RENDER -->
+```{r analysis, include=FALSE}
+source(here::here("R/analysis.R"))
+```
 
@@
 
-![Field sampling locations](images/fig1_map.jpg){#fig-map width=100%}
+![Field sampling locations](images/fig1_map.jpg){#fig-map width="6in" height="8in"}
 
@@
 
-![Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg){#fig-chlascatter width=80%}
+![Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg){#fig-chlascatter width="8.5in" height="5.25in"}
 
@@
 
-![Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width=80%}
+![Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width="8.5in" height="5.25in"}
 
@@
 
-![Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){#fig-relbv width=80%}
+![Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){#fig-relbv width="8.5in" height="5.25in"}
 
-![Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg){#fig-bv width=80%}
+![Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg){#fig-bv width="8.5in" height="5.25in"}
 
@@
 
-![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg){#fig-chlcount width=80%}
+![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
 
-![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg){#fig-phycocount width=80%}
+![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
 

### 2023-12-04 manuscript/manuscript.qmd
starting on summary stuff

@@
 bibliography: manuscript.bib
-#filters: 
-#  - authors-block
 ---
@@
 
-To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots to assess the fit and compare the different sensors.
-We derived the coefficient of determination from the regressions as a measure of fit and qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts.
-We do not compare slopes as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.
+To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots.
+We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts with scatterplots.
+We do not compare slopes nor look for a slope of 1 as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.
 
@@
 Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran.
-Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in-vivo concentrations of chlorophyll (Zamyadi et al .
+Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (Zamyadi et al .
 2012).
@@
 
-This study supports the use of extracted chlorophll and phycocyanin as an alternative to cell counts for the study waterbodies.
+This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies.
 Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments.

### 2023-12-04 R/functions.R

@@ -630,3 +630,8 @@ map_field_sites <- function(){
   sites_gg
   
 }
+
+summarize_data <- function(fluoro_df, phyco_df){
+  browser()
+  
+}

### 2023-12-04 manuscript/manuscript.qmd
Jeff edits on paper.  Ready for Betty review?!?

@@
 We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.
-We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (r2 ranging from 0.78-0.96).
-We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (r2 between 0.34 and 0.82).
+We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (R^2^ ranging from 0.78-0.96).
+We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).
 Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.
@@
 Differences in measurement methods represent a potential source of error when comparing instruments.
-Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
-In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light.
+Additionally, instruments may report valued in RFUs (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of RFUs as photosystems become saturated with light.
 Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior.
@@
 
-## Summary of Data
-
-<!-- Jeff -->
-
-Need summary of sites and dataFig 1. Map of Field Sites
-
 ## Chlorophyll
@@
 Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond).
-Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
+Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R^2^ values of 0.82 and 0.81, respectively.  The Cyanofluor (R^2^=0.35) and Trilogy in vivo (R^2^=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
 
@@
 Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond).
-The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter).
-Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin.
+The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter).
+Phycocyanin measurements for the Fluorosense (R^2^=0.96) and the Cyanofluor (R^2^=0.78) were also strongly related with extracted phycocyanin.
 
@@
 
-Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59).
-The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount).
-Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(@fig-phycocount).
+Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (R^2^=0.92) followed by the algaetorch (R^2^=0.78) and the Phycoprobe (R^2^=0.59).
+The Cyanofluor (R^2^=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount).
+Phycocyanin measurements from all instruments were strongly correlated with cell counts (R^2^ between 0.9 and 0.97)(@fig-phycocount).
 
@@
 The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations but the output from these instruments is not a direct measure of phycocyanin; instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria.
-The Cyanofluor output was in rfus and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study.
+The Cyanofluor output was in RFUs and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study.
 A study by Thomson-Laing et al.
-did convert rfus to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (Thomson-Laing et al.
+did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (Thomson-Laing et al.
 2020).
@@
 
-All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll a.
+All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll *a*.
 All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship.
@@
 
-# Summary
-
-1.) In general, in situ phycocyanin better match to extracted than in situ chlorophyll is to extracted. The cholorophylll relationship is driven by phytoplankton community make up
-
-2.) Extracted chlorophyll and phycocyanin, are great match to cell counts. Suggests pigments are a good proxy for cell counts and could be used in place of cell counts.
-
-3.) A discussion with water quality managers about using extracted or in situ phycocyanin as a replacement for cell counts might be warranted.
-
-Don't read anything into absolute values (i.e.
-chlorophyll from one sensor not equal to chlorophyll from another sensor) across sensors, but general patterns are still very useful.
-Need to know info about specific sites to be useful.
-
-Phycocyanin very good proxy for cell counts.
-Suggest replacing cell counts for waterbody listing decisions with phycocyanin.
-Would likely need additional work to furhter explore relationship, build predictive models of cell coutns, etc.
-
 # Acknowledgements
 
-And special thanks to Gracie, Anastasia, and Lilac for emotional support during challenging analysis times.
+We are grateful to EPA Reviewer 1, EPA Reviewer 2, EPA Reviewer 3, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper.
+The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.
+Environmental Protection Agency.
+Any mention of trade names, products, or services does not imply an endorsement by the U.S.
+Government or the U.S.
+Environmental Protection Agency.
+The EPA does not endorse any commercial products, services, or enterprises.
+This contribution is identified by the tracking number ORD-XXXXXX of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency.
+Lastly, special thanks to Gracie, Anastasia, and Lilac for cheer leading and support during challenging analysis times.
 

### 2023-12-04 R/analysis.R

@@ -76,3 +76,4 @@ ggsave(here::here("manuscript/images/fig7_phyco_vs_cells.jpg"), fluoro_phyco_cou
 field_sites <- map_field_sites() 
 ggsave(here::here("manuscript/images/fig1_map.jpg"), field_sites, 
        width = 6, height = 8, dpi = 300)
+

### 2023-12-04 R/functions.R

@@ -630,8 +630,3 @@ map_field_sites <- function(){
   sites_gg
   
 }
-
-summarize_data <- function(fluoro_df, phyco_df){
-  browser()
-  
-}

### 2023-12-04 manuscript/manuscript.qmd
added summary tables

@@
 
+<!-- Need summary table of chl and phyco for each waterbody/instrument-->
+
 ## Chlorophyll
 
-Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond).
+Extracted chlorophyll measurements ranged from 3.2 ug/L at Indian Lake to 81.3 ug/L in Mashapaug Pond (@tbl-chlsummary).
 Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R^2^ values of 0.82 and 0.81, respectively.  The Cyanofluor (R^2^=0.35) and Trilogy in vivo (R^2^=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
@@
 
+```{r chl_summary, results='asis', echo=FALSE, message=FALSE}
+#| label: tbl-chlsummary
+#| tbl-cap: "Chlorophyll summary data by waterbody and instrument."
+
+knitr::kable(fluoro_summary$chl_summary)
+```
+
 ## Phycocyanin
 
-Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond).
+Extracted phycocyanin measurements ranged from 1.3 ug/L at Indian Lake to 75.0 ug/L in Mashapaug Pond (@tbl-phycosummary).
 The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter).
@@
 ![Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width="8.5in" height="5.25in"}
+
+```{r phyco_summary, results='asis', echo=FALSE, message=FALSE}
+#| label: tbl-phycosummary
+#| tbl-cap: "Phycocyanin summary data by waterbody and instrument."
+
+knitr::kable(fluoro_summary$phyco_summary)
+```
 
@@
 
+```{r session_info, echo=FALSE}
+#Send session info to a file: 
+capture.output(sessionInfo(),file=here::here("sessioninfo.txt"))
+```
+
 # References

### 2023-12-04 R/analysis.R

@@ -77,3 +77,5 @@ field_sites <- map_field_sites()
 ggsave(here::here("manuscript/images/fig1_map.jpg"), field_sites, 
        width = 6, height = 8, dpi = 300)
 
+# Fluoro summary
+fluoro_summary <- summary_table(fp_data_wb)

### 2023-12-04 R/functions.R

@@ -630,3 +630,14 @@ map_field_sites <- function(){
   sites_gg
   
 }
+
+summary_table <- function(fluoro_df){
+  idx <- !(fluoro_df$instrument == "trilogy" & fluoro_df$units == "rfu")
+  summ_df <- fluoro_df |>
+    filter(idx) |>
+    filter(variable %in% c("chl", "phyco")) |>
+    group_by(waterbody, instrument, variable, units) |>
+    summarise(mean = mean(avg_value, na.rm = TRUE), sd = sd(avg_value, na.rm = TRUE))
+  list(chl_summary = summ_df[summ_df$variable == "chl",], 
+       phyco_summary = summ_df[summ_df$variable == "phyco",])
+}

### 2023-12-08 manuscript/manuscript.qmd
added references

--- 2023-12-04 manuscript/manuscript.qmd
@@
 bibliography: manuscript.bib
+#filters: 
+#  - authors-block
 ---
@@
 
-Harmful algal blooms caused by cyanobacteria (cyanoHABs) are likely to increase in a warming world with greater inputs of nutrients by humans.
+Harmful algal blooms caused by cyanobacteria (cyanoHABs) are likely to increase in a warming world with greater inputs of nutrients by humans (Paerl et al.
+2016; Schindler et al.
+2008).
 These blooms may impact both human and environmental health, particularly if toxins are being produced.
-CyanoHAB blooms are often transient in nature, can be difficult to accurately quantify, and occur in diverse waterbodies across large spatial scales.
-Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions.
+CyanoHAB blooms are often transient in nature, can be difficult to accurately quantify, and occur in diverse waterbodies across large spatial scales (Wynne et al.
+2010).
+Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions (Chaffin et al.
+2018; Gobler et al.
+2016).
 As large, centrally managed monitoring efforts are expensive and time consuming to operate, a substantial portion of cyanoHAB monitoring occurs at the local level.
@@
 Some instruments (i.e.
-PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition.
+PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (Buetler et al.
+2002).
 These instruments can be used in the field, the lab, or a combination of both.
@@
 Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
-In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light.
-Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior.
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (Liu and Georgakakos 2021).
+Estimation problems can also result from differing community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (Rousso et al.
+2022; .
 Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
-Some instruments offer a yellow substance correction to help account for CDOM while others do not.
+Some instruments offer a yellow substance correction to help account for CDOM while others do not (Cremella et al.
+2018).
 How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types.
@@
 
-To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots.
-We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts with scatterplots.
-We do not compare slopes nor look for a slope of 1 as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.
+To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots to assess the fit and compare the different sensors.
+We derived the coefficient of determination from the regressions as a measure of fit and qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts.
+We do not compare slopes as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.
 
@@
 Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran.
-Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (Zamyadi et al .
+Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in-vivo concentrations of chlorophyll (Zamyadi et al .
 2012).
@@
 
-This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies.
+This study supports the use of extracted chlorophll and phycocyanin as an alternative to cell counts for the study waterbodies.
 Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments.

### 2023-12-19 manuscript/manuscript.qmd

--- 2023-12-04 manuscript/manuscript.qmd
@@
 Differences in measurement methods represent a potential source of error when comparing instruments.
-Additionally, instruments may report valued in RFUs (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
-In vivo fluorescence can be affected by photochemical quenching causing an underestimation of RFUs as photosystems become saturated with light.
-Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior.
+Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (Liu and Georgakakos 2021).
+Estimation problems can also result from differing community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (Rousso et al.
+2022; .
 Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
-Some instruments offer a yellow substance correction to help account for CDOM while others do not.
+Some instruments offer a yellow substance correction to help account for CDOM while others do not (Cremella et al.
+2018).
 How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types.

### 2023-12-19 manuscript/manuscript.qmd
example citations

@@
 2010).
-Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions (Chaffin et al.
-2018; Gobler et al.
-2016).
+Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions [@chaffin_interactions_2018; @gobler_dual_2016].
 As large, centrally managed monitoring efforts are expensive and time consuming to operate, a substantial portion of cyanoHAB monitoring occurs at the local level.

### 2023-12-21 manuscript/manuscript.qmd
bibliography added to manuscript

@@
 
-Harmful algal blooms caused by cyanobacteria (cyanoHABs) are likely to increase in a warming world with greater inputs of nutrients by humans (Paerl et al.
-2016; Schindler et al.
-2008).
+Harmful algal blooms caused by cyanobacteria (cyanoHABs) are likely to increase in a warming world with greater inputs of nutrients by humans (@paerl_it_2016; @schindler_eutrophication_2008).
 These blooms may impact both human and environmental health, particularly if toxins are being produced.
-CyanoHAB blooms are often transient in nature, can be difficult to accurately quantify, and occur in diverse waterbodies across large spatial scales (Wynne et al.
-2010).
+CyanoHAB blooms are often transient in nature, can be difficult to accurately quantify, and occur in diverse waterbodies across large spatial scales (@wynne_characterizing_2010).
 Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions [@chaffin_interactions_2018; @gobler_dual_2016].
@@
 Some instruments (i.e.
-PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (Buetler et al.
-2002).
+PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002).
 These instruments can be used in the field, the lab, or a combination of both.
@@
 Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
-In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (Liu and Georgakakos 2021).
-Estimation problems can also result from differing community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (Rousso et al.
-2022; .
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (@liu_chlorophyll_2021).
+Estimation problems can also result from differing community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022).
 Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
-Some instruments offer a yellow substance correction to help account for CDOM while others do not (Cremella et al.
-2018).
+Some instruments offer a yellow substance correction to help account for CDOM while others do not (@cremella_interpretation_2018).
 How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types.
@@
 Hambdhani et al.
-also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations (Hambdhani et al.
-2021).
+also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations (@hamdhani_performance_2021).
 Therefore, using this particular instrument to enforce set criteria across a variety of waterbodies would probably not be recommended without further research.
@@
 A study by Thomson-Laing et al.
-did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (Thomson-Laing et al.
-2020).
+did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (@thomson-laing_predicting_2020).
 They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.
@@
 Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran.
-Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (Zamyadi et al .
-2012).
+Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012).
 Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments.
 A study by Silva et al.
-found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al.
-2016).
+found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (@silva_comparison_2016).
 

### 2024-01-04 manuscript/manuscript.qmd
updated intro with Betty's comments

@@
 ---
-title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
+title: "Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
 author: 
@@
 A variety of instruments have been developed to measure cyanoHAB indicators (i.e.
-chlorophyll *a* or phycocyanin) that do not require an extraction process.
-We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.
-We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (R^2^ ranging from 0.78-0.96).
-We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).
+chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g.
+pigment extraction).
+We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.
+We found a strong relationship between extracted phycocyanin and  in vivo fluorometers (R^2^ ranging from 0.78-0.96).
+We found less consistency between  in vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).
 Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.
-Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.
+Phycocyanin results from  in vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.
 Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.
 Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.
-The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
+The results from this study encourage further research using in vivo fluorometry and phycocyanin for cyanoHAB monitoring efforts and when making decisions regarding public health."
 format: 
@@
 
-A variety of instruments have been developed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step and have a wide range in cost.
+A variety of instruments have been developed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step.
 These instruments distinguish between the pigments based on their differing emission spectra.
@@
 PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002).
-These instruments can be used in the field, the lab, or a combination of both.
+These instruments can be used in the field, the lab, a combination of both, and have a wide range in cost.
 
@@
 
-Field instruments may be placed directly in a waterbody or require water to be poured into a cuvette and then placed into a handheld fluorometer.
+Field instruments may be placed directly in a waterbody or require sample water to be poured into a cuvette and then placed into a handheld fluorometer.
 Differences in measurement methods represent a potential source of error when comparing instruments.
-Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
+Additionally, instruments may report values in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
 In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (@liu_chlorophyll_2021).
-Estimation problems can also result from differing community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022).
-Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
+Estimation problems can also result from differing phytoplankton/cyanobacteria community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022).
+Additionally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
 Some instruments offer a yellow substance correction to help account for CDOM while others do not (@cremella_interpretation_2018).
-How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types.
-The goal of this paper is to compare different handhelds or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.
+
+Overall, the physical properties and phytoplankton community composition of a waterbody can cause instruments to estimate pigment concentrations differently.
+How the different instruments handle these various potential situations can affect their accuracy and precision, and the ability to compare across instrument types.
+The goal of this paper is to compare several handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.
 
@@
 Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
-Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in-vivo module.
+Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module.
 chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
@@
 
-The results from this study indicate that in-vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll.
+The results from this study indicate that in vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll.
 Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements.
@@
 Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts.
-There are a variety of instruments capable of measuring phycocyanin using in-vivo fluorometry across a spectrum of prices.
-We hope this study will encourage further research and consideration of using in-vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.
+There are a variety of instruments capable of measuring phycocyanin using in vivo fluorometry across a spectrum of prices.
+We hope this study will encourage further research and consideration of using in vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.
 

### 2024-01-05 manuscript/manuscript.qmd
updated methods with details on handling of field/lab replicates

@@
 Field samples were collected from eight ponds in Rhode Island during October and November 2021 (@fig-map).
-Two liter surface samples were collected in triplicate (six total liters) by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment.
+Surface samples were collected in triplicate by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment.
 If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles.
@@
 
-Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
+Within 24 hours of collection, three replicates (i.e.
+lab replicates) from each field sample was analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
 Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module.
@@
 All phytoplankton samples were collected on 2021-10-06.
-Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22; fluoromery was performed with 24 hours of sample collection.
+Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22; fluorometry was performed with 24 hours of sample collection.
 
@@
 
+<!-- Clean up and add details on replicate/duplicate averaging -->
+
 To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots.
+For these regressions and scatterplots, the lab replicates were averaged, resulting in three field replicates for each of the waterbodies.
 We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts with scatterplots.

### 2024-01-12 manuscript/manuscript.qmd
finished edits based on Betty's comments

@@
 
-The following five fluorometers were used during this study: Turner Trilogy (in vivo chlorophyll module), Turner FluoroSense, Turner Cyanofluor, bbe AlgaeTorch, and bbe PhycoProbe.
+The following five fluorometers (in decreasing order of cost) were used during this study: bbe PhycoProbe, bbe AlgaeTorch, Turner Trilogy (in vivo chlorophyll and orange modules), Turner Cyanofluor, and Turner FluoroSense.
+These fluorometers were chosen based on their availability to us as either our lab or our collaborators in the northeast US have used them.
 Measurements from these fluorometers were compared with the results of a solvent-based extraction for chlorophyll and phycocyanin using a Turner Trilogy with a chla-na and orange module, respectively.
@@
 Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module.
-chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
+Chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
 Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer.
-Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).
+Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin)(\@kasinak2015benchtop; \@welschmeyer1994fluorometric).
 
@@
 Samples were stored at 4°C and shipped the week after collection on ice to Phycotech, Inc.
+A full description of methods and a quality assurance plan are available from Phycotech, Inc.
+(https://www.phycotech.com/Portals/0/PDFs/PhycoTech_INFO_PACKET.pdf).
 
@@
 Extracted phycocyanin measurements ranged from 1.3 ug/L at Indian Lake to 75.0 ug/L in Mashapaug Pond (@tbl-phycosummary).
-The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter).
+The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations (@fig-phycoscatter).
 Phycocyanin measurements for the Fluorosense (R^2^=0.96) and the Cyanofluor (R^2^=0.78) were also strongly related with extracted phycocyanin.
@@
 
-The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations but the output from these instruments is not a direct measure of phycocyanin; instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria.
+The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations.
+However, the output from these instruments is not a direct measure of phycocyanin.
+Instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria, which makes a direct comparision with other instruments that measure phycocyanin difficult.
 The Cyanofluor output was in RFUs and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study.

### 2024-02-20 manuscript/manuscript.qmd
switched to cells/ml on Fig 4

@@
 
-Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-relbv).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(@fig-bv).
+!!!REWRITE ME - Changed from biov to cells/ml, only one fig now!!!!
+
+Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-cellsml).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(@fig-bv).
 Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).
 
-![Compare Relative Biovolume of divisions by waterbody](images/fig4_rel_bio_bar.jpg){#fig-relbv width="8.5in" height="5.25in"}
-
-![Compare Total Biovlume of divisions by waterbody](images/fig5_total_bio_bar.jpg){#fig-bv width="8.5in" height="5.25in"}
+![Compare Cells/mL of divisions by waterbody](images/fig4_total_cellsml_bar.jpg){#fig-cellsml width="8.5in" height="5.25in"}
 
@@
 
-![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig6_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
+![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig5_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
 
-![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig7_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
+![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig6_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
 

### 2024-02-20 R/functions.R

@@ -502,7 +502,7 @@ clean_phycotech_cyano <- function(phycotech_df){
 
 #' Make grouped bar plot
 grouped_bar_plot <- function(phycotech_df, yvar){
-  
+ 
   yvar <- rlang::sym(yvar)
   bar_plot <- phycotech_df %>%
     ggplot(aes(x = waterbody, y = !!yvar, fill = barplot_groups)) +
@@ -637,7 +637,12 @@ summary_table <- function(fluoro_df){
     filter(idx) |>
     filter(variable %in% c("chl", "phyco")) |>
     group_by(waterbody, instrument, variable, units) |>
-    summarise(mean = mean(avg_value, na.rm = TRUE), sd = sd(avg_value, na.rm = TRUE))
+    summarise(mean = mean(avg_value, na.rm = TRUE), sd = sd(avg_value, na.rm = TRUE)) |>
+    ungroup()
+  summ_df <- filter(summ_df, instrument == "trilogy") |>
+    mutate(mean = round(mean, 1), sd = round(sd, 1)) |>
+    select(!instrument) |>
+    as.data.frame()
   list(chl_summary = summ_df[summ_df$variable == "chl",], 
        phyco_summary = summ_df[summ_df$variable == "phyco",])
 }

### 2024-02-20 R/functions.R
added cell counts to supplemental table

@@ -481,7 +481,9 @@ clean_phycotech_cyano <- function(phycotech_df){
   phycotech_df_clean <- phycotech_df %>%
     select(waterbody = system_name, date = sample_date, division, family, genus, 
            biovolume_concentration = total_biovolume_cubic_um_per_ml_,
-           relative_biovolume = relative_total_biovolume) %>%
+           relative_biovolume = relative_total_biovolume,
+           concentration = concentration_natural_units_per_ml_,
+           relative_concentration) %>%
     # TODO: Standardize names with our data
     mutate(date = ymd(date),
            waterbody = case_when(waterbody == "Yawagoo Pond" ~

### 2024-02-20 manuscript/manuscript_old.qmd
made edits but have no idea what i did

--- 2024-01-12 manuscript/manuscript.qmd
@@
 Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer.
-Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin)(\@kasinak2015benchtop; \@welschmeyer1994fluorometric).
+Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin)(@kasinak2015benchtop; @welschmeyer1994fluorometric).
 

### 2024-04-05 manuscript/manuscript.qmd
edits made to qmd

@@
 Despite occuring on large scales, effects of blooms have primarily local impacts and can differ in species composition and toxicity based on local conditions [@chaffin_interactions_2018; @gobler_dual_2016].
-As large, centrally managed monitoring efforts are expensive and time consuming to operate, a substantial portion of cyanoHAB monitoring occurs at the local level.
+As large, centrally managed monitoring efforts are expensive and time consuming to operate, a substantial portion of cyanoHAB monitoring occurs on a smaller, more local level.
 As neighborhood and lake associations generally do not have the equipment or training to perform extractive analysis of cholorophyll a and/or phycocyanin (pigments found in all algae and specifically cyanobacteria, respectively), alternative measurement equipment has been developed.
@@
 
-A variety of instruments have been developed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step.
+A variety of instruments have been designed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step.
 These instruments distinguish between the pigments based on their differing emission spectra.
 Some instruments (i.e.
-PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002).
+PhycoProbe) use multiple leds at varying wavelengths to cause excitation and emission of different algal type pigments, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002).
 These instruments can be used in the field, the lab, a combination of both, and have a wide range in cost.
@@
 Estimation problems can also result from differing phytoplankton/cyanobacteria community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022).
-Additionally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM.
+Additionally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of colored dissolved organic matter (CDOM).
 Some instruments offer a yellow substance correction to help account for CDOM while others do not (@cremella_interpretation_2018).
@@
 How the different instruments handle these various potential situations can affect their accuracy and precision, and the ability to compare across instrument types.
-The goal of this paper is to compare several handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.
+The goal of this paper is to compare several handheld and in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.
 
@@
 The following five fluorometers (in decreasing order of cost) were used during this study: bbe PhycoProbe, bbe AlgaeTorch, Turner Trilogy (in vivo chlorophyll and orange modules), Turner Cyanofluor, and Turner FluoroSense.
-These fluorometers were chosen based on their availability to us as either our lab or our collaborators in the northeast US have used them.
+These fluorometers were chosen based on their availability to us as many have been used by our collaborators in the northeast US.
 Measurements from these fluorometers were compared with the results of a solvent-based extraction for chlorophyll and phycocyanin using a Turner Trilogy with a chla-na and orange module, respectively.
@@
 If a surface scum was present, the scum was gently brushed aside to avoid being collected in the sample bottles.
-Samples were collected in acid-washed 1 L amber bottles and placed in a cooler with ice.
+Samples were collected in acid-washed 1 L amber bottles, placed in a cooler with ice, and stored at 4°C until analysis.
 At the same location as sample collection, measurements of chlorophyll and phycocyanin were made using the AlgaeTorch and the FluoroSense.
@@
 Within 24 hours of collection, three replicates (i.e.
-lab replicates) from each field sample was analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
+lab replicates to assess precision) from each field sample were analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
 Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module.
-Chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath.
-Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer.
-Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin)(\@kasinak2015benchtop; \@welschmeyer1994fluorometric).
+Chlorophyll *a* was extracted by placing frozen GF/F filters in centrifuge tubes with 90% acetone, then placing the tubes for a 20 minute period in a sonicating water bath to lyse cyanobacterial cells (\@welschmeyer1994fluorometric).
+Phycocyanin was extracted from frozen GF/F filters using a 50 mM phosphate buffer as the extraction solvent and following a 15 minute period in a sonicating water bath to lyse cyanobacterial cells (\@kasinak2015benchtop).
+Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).
 
@@
 Samples were stored at 4°C and shipped the week after collection on ice to Phycotech, Inc.
+A minimum of 400 natural units were counted per sample, and quantitative biovolume estimates were also performed for each sample.
 A full description of methods and a quality assurance plan are available from Phycotech, Inc.
@@
 
-To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots.
+To compare extracted concentrations and cell counts to the various instruments we used simple linear regression and scatterplots.
 For these regressions and scatterplots, the lab replicates were averaged, resulting in three field replicates for each of the waterbodies.
@@
 
-!!!REWRITE ME - Changed from biov to cells/ml, only one fig now!!!!
+!!! REWRITE ME - Changed from biov to cells/ml, only one fig now !!!! !!! Cyanobaceteria Cell Counts are natural units / mL !!!!
 

### 2024-04-05 manuscript/manuscript.qmd

@@
   eval = TRUE,
-  cache = TRUE,
   tidy = FALSE,
@@
 
-```{r analysis, include=FALSE}
+```{r analysis, include=FALSE, cache=FALSE}
 source(here::here("R/analysis.R")) 
@@
 #| label: tbl-chlsummary
-#| tbl-cap: "Chlorophyll summary data by waterbody and instrument."
+#| tbl-cap: "Extracted chlorophyll summary data by waterbody and instrument."
 
@@
 #| label: tbl-phycosummary
-#| tbl-cap: "Phycocyanin summary data by waterbody and instrument."
+#| tbl-cap: "Extracted phycocyanin summary data by waterbody and instrument."
 
@@
 
-Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-cellsml).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(@fig-bv).
+<!-- phycotech data is in data/phycotech_data_summ.csv" -->
+
+Cyanophytes were the dominant taxa by natural units / mL at Curran (???), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-cellsml).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^).
 Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).

### 2024-04-12 manuscript/manuscript.qmd
finished edits from internal reviews

@@
 Differences in measurement methods represent a potential source of error when comparing instruments.
-Additionally, instruments may report values in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
-In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (@liu_chlorophyll_2021).
+Additionally, instruments may report values in RFUs (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison.
+In vivo fluorescence can be affected by photochemical quenching causing an underestimation of RFUs as photosystems become saturated with light (@liu_chlorophyll_2021).
 Estimation problems can also result from differing phytoplankton/cyanobacteria community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022).
@@
 
-Cyanophytes were the dominant taxa by natural units / mL at Curran (???), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-cellsml).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^).
+Cyanophytes were the dominant taxa at all waterbodies with the highest concentrations observed at Curran (38,019 natural units/mL), Mashapaug (89,806 natural units/mL), and Warwick (45,847 natural units/mL)(@fig-cellsml).  Total concentrations were highest at Mashapaug (92,527 natural units/mL), Warwick (70,728 natural units/mL), and Curran (40,428 natural units/mL).
 Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).
@@
 Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements.
-Instruments measuring phycocyanin were better able to handle differences in the phytoplankton community between these study ponds that cause chlorophyll measurements to be less strongly correlated with extracted results.
+Instruments measuring in-vivo phycocyanin were better able to handle the cyanobacteria-dominated phytoplankton communities in these study ponds that cause in-vivo chlorophyll measurements to be less strongly correlated with extracted results.
 Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.
@@
 The relationship between phycocyanin and all fluorometers was strong.
-The Fluorosense had the strongest correlation even with the differences in phytoplankton concentrations and community dynamics between the waterbodies.
-However, the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L).
+The Fluorosense had the strongest correlation between the waterbodies, however the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L).
 Hambdhani et al.
@@
 
-The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations.
+The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations compared to extracted concentrations.
 However, the output from these instruments is not a direct measure of phycocyanin.
-Instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria, which makes a direct comparision with other instruments that measure phycocyanin difficult.
+Instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria, which makes a comparision with other instruments that directly measure phycocyanin difficult.
 The Cyanofluor output was in RFUs and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study.
@@
 Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran.
-Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012).
+Phycocyanin concentrations and cell counts of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012).
 Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments.

### 2024-04-15 manuscript/manuscript.qmd
continued edits

@@
 
-Field samples were collected from eight ponds in Rhode Island during October and November 2021 (@fig-map).
+Field samples were collected from six ponds in Rhode Island.
+(@fig-map).
+Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22.
 Surface samples were collected in triplicate by wading into each pond to a minimum depth of two feet to avoid collecting near the sediment.
@@
 
+All phytoplankton samples were collected on 2021-10-06.
 Water samples for phytoplankton identification and enumeration were decanted from field collected samples into 250 ml amber HDPE bottles and were preserved with 25% glutaraldehyde.
@@
 (https://www.phycotech.com/Portals/0/PDFs/PhycoTech_INFO_PACKET.pdf).
-
-All phytoplankton samples were collected on 2021-10-06.
-Samples from Barber, Indian and Yawgoo were collected on 2021-10-06 and samples from Curran, Mashapaug, and Warwick were collected on 2021-10-22; fluorometry was performed with 24 hours of sample collection.
 
@@
 For these regressions and scatterplots, the lab replicates were averaged, resulting in three field replicates for each of the waterbodies.
-We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts with scatterplots.
+We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compared the device measurements to the extracted concentrations and phytoplankton counts with scatterplots.
 We do not compare slopes nor look for a slope of 1 as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.
@@
 
-Extracted chlorophyll measurements ranged from 3.2 ug/L at Indian Lake to 81.3 ug/L in Mashapaug Pond (@tbl-chlsummary).
+Extracted chlorophyll measurements ranged from a mean of 1.3 ug/L in Indian Lake to a mean of 32.2 ug/L in Mashapaug Pond (@tbl-chlsummary).
 Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R^2^ values of 0.82 and 0.81, respectively.  The Cyanofluor (R^2^=0.35) and Trilogy in vivo (R^2^=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.
@@
 
-Extracted phycocyanin measurements ranged from 1.3 ug/L at Indian Lake to 75.0 ug/L in Mashapaug Pond (@tbl-phycosummary).
+Extracted phycocyanin measurements ranged from a mean of 0.2 ug/L in Indian Lake and Barber Pond to 10.5 ug/L in Mashapaug Pond (@tbl-phycosummary).
 The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations (@fig-phycoscatter).
@@
 
-Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (R^2^=0.92) followed by the algaetorch (R^2^=0.78) and the Phycoprobe (R^2^=0.59).
+Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (R^2^=0.92) followed by the Algaetorch (R^2^=0.78) and the Phycoprobe (R^2^=0.59).
 The Cyanofluor (R^2^=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount).
@@
 
-We are grateful to EPA Reviewer 1, EPA Reviewer 2, EPA Reviewer 3, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper.
+We are grateful to Hilary Snook, Darryl Keith, Steve Rego, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper.
 The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.

### 2024-04-18 manuscript/manuscript.qmd

@@
 
-The following five fluorometers (in decreasing order of cost) were used during this study: bbe PhycoProbe, bbe AlgaeTorch, Turner Trilogy (in vivo chlorophyll and orange modules), Turner Cyanofluor, and Turner FluoroSense.
-These fluorometers were chosen based on their availability to us as many have been used by our collaborators in the northeast US.
+For this study, we used five fluorometers that are commonly used by our collaborators in the northeast US.
+In decreasing order of cost, they are: bbe PhycoProbe, bbe AlgaeTorch, Turner Trilogy (in vivo chlorophyll and orange modules), Turner Cyanofluor, and Turner FluoroSense.
 Measurements from these fluorometers were compared with the results of a solvent-based extraction for chlorophyll and phycocyanin using a Turner Trilogy with a chla-na and orange module, respectively.

### 2024-07-19 manuscript/manuscript.qmd
Fresh render
MC revisions for BC STICS review

@@
 
-![Comparison of fluorometers by waterbody](images/fig2_chla_scatter.jpg){#fig-chlascatter width="8.5in" height="5.25in"}
+![Comparison of fluorometer measurements of chlorophyll by waterbody.
+Blue lines represent the best fit regression line.
+R^2^ on plot is the coeffecient of determination for this line.](images/fig2_chla_scatter.jpg){#fig-chlascatter width="8.5in" height="5.25in"}
 
@@
 #| label: tbl-chlsummary
-#| tbl-cap: "Extracted chlorophyll summary data by waterbody and instrument."
+#| tbl-cap: "Summary of extracted chlorophyll from Turner Trilogy by waterbody."
 
-knitr::kable(fluoro_summary$chl_summary)
+knitr::kable(fluoro_summary$chl_summary, row.names = FALSE)
 ```
@@
 
-![Comparison of phycocyanin measurements by waterbody](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width="8.5in" height="5.25in"}
+![Comparison of fluorometer measurements of phycocyanin by waterbody.
+Blue lines represent the best fit regression line.
+R^2^ on plot is the coeffecient of determination for this line.](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width="8.5in" height="5.25in"}
 
@@
 #| label: tbl-phycosummary
-#| tbl-cap: "Extracted phycocyanin summary data by waterbody and instrument."
+#| tbl-cap: "Summary of extracted phycocyanin from Turner Trilogy by waterbody."
 
-knitr::kable(fluoro_summary$phyco_summary)
+knitr::kable(fluoro_summary$phyco_summary, row.names = FALSE)
 ```
@@
 ## Phytoplankton Counts
-
-!!! REWRITE ME - Changed from biov to cells/ml, only one fig now !!!! !!! Cyanobaceteria Cell Counts are natural units / mL !!!!
 
@@
 
-![Compare fluorometer chlorophyll to cyanobacteria cell counts](images/fig5_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
+![Compare fluorometer chlorophyll to cyanobacteria cell counts.
+Blue lines represent the best fit regression line.
+R^2^ on plot is the coeffecient of determination for this line.](images/fig5_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
 
-![Compare fluorometer phycocyanin to cyanobacteria cell counts](images/fig6_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
+![Compare fluorometer phycocyanin to cyanobacteria cell counts.
+Blue lines represent the best fit regression line.
+R^2^ on plot is the coeffecient of determination for this line.](images/fig6_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
 
@@
     mutate(biovolume = round(biovolume, 1),
-         relative_biovolume = round(relative_biovolume, 5))
+           relative_biovolume = round(relative_biovolume, 5)) %>%
+    filter(waterbody %in% c("Barber Pond", "JL Curran Reservoir", "Indian Lake", "Mashapaug Pond", "Warwick Pond", "Yawgoo"))
 ```
@@
 #| label: tbl-supp
-#| tbl-cap: "Waterbody, Cyano Genus, Relative Biovolume Total Biovolume"
+#| tbl-cap: "Cyanobacteria community composition by genus and waterbody, including total biovolume and relative biovolume."
 
-knitr::kable(table1)
+knitr::kable(table1, row.names = FALSE)
 ```

### 2024-07-19 R/functions.R

@@ -302,7 +302,7 @@ ext_vs_all_plot <- function(fpdata, var, meth, x_order = NULL){
   myplot <- plot_data %>%
     ggplot(aes(x = extracted_value, y = avg_value)) +
     geom_point(size = 4, aes(color = waterbody)) +
-    geom_smooth(data = plot_data, method = "lm") +
+    geom_smooth(data = plot_data, method = "lm", se = FALSE) +
     geom_blank(data = blank_data) +
     facet_wrap(instrument_unit ~ ., scales = "free", strip.position = "left") +
     theme_ipsum_rc() +
@@ -587,7 +587,7 @@ flouro_vs_count_plot <- function(fluoro_df, phycotech_df, xvar = c("chlorophyll"
   myplot <- plot_df %>%
     ggplot(aes(x = concentration, y = yvar)) +
     geom_point(size = 4, aes(color = waterbody)) +
-    geom_smooth(method = "lm") +
+    geom_smooth(method = "lm", se = FALSE) +
     facet_wrap(instrument ~ ., scales = "free", strip.position = "left") +
     theme_ipsum_rc() +
     scale_color_viridis_d(option = "plasma") +

### 2024-07-19 manuscript/manuscript.qmd
edits based on feedback from MC (BC review)

@@
 abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.
-A variety of instruments have been developed to measure cyanoHAB indicators (i.e.
-chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g.
-pigment extraction).
+A variety of instruments have been developed to measure cyanoHAB indicators (i.e., chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g., pigment extraction).
 We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.
@@
 These instruments distinguish between the pigments based on their differing emission spectra.
-Some instruments (i.e.
-PhycoProbe) use multiple leds at varying wavelengths to cause excitation and emission of different algal type pigments, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002).
+Some instruments (i.e., PhycoProbe) use multiple leds at varying wavelengths to cause excitation and emission of different algal type pigments, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002).
 These instruments can be used in the field, the lab, a combination of both, and have a wide range in cost.
@@
 
-![Field sampling locations](images/fig1_map.jpg){#fig-map width="6in" height="8in"}
+![Six field sampling locations in Rhode Island, USA](images/fig1_map.jpg){#fig-map width="6in" height="8in"}
 
@@
 
-Within 24 hours of collection, three replicates (i.e.
-lab replicates to assess precision) from each field sample were analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
+Within 24 hours of collection, three replicates (i.e., lab replicates to assess precision) from each field sample were analyzed using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction.
 Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module.
@@
 
-![Compare Cells/mL of divisions by waterbody](images/fig4_total_cellsml_bar.jpg){#fig-cellsml width="8.5in" height="5.25in"}
+![Cell counts (cells/mL) of phytoplankton samples from six waterbodies in Rhode Island](images/fig4_total_cellsml_bar.jpg){#fig-cellsml width="8.5in" height="5.25in"}
 
@@
 The results from this study indicate that in vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll.
-Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements.
+Similarly, phycocyanin from all instruments was better correlated to cell counts than correlations between chlorophyll measurements from instruments and cell counts.
 Instruments measuring in-vivo phycocyanin were better able to handle the cyanobacteria-dominated phytoplankton communities in these study ponds that cause in-vivo chlorophyll measurements to be less strongly correlated with extracted results.
-Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.
+Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds, suggesting that pigments are indeed a good proxy for cell counts, at least for some waterbodies.
 
 The relationship between phycocyanin and all fluorometers was strong.
-The Fluorosense had the strongest correlation between the waterbodies, however the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L).
+The Fluorosense had the strongest correlation between the waterbodies; however, the Fluorosense overestimated the actual concentrations as compared to extracted values (0-35 ug/L vs 0-12.5 ug/L).
 Hambdhani et al.
-also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations (@hamdhani_performance_2021).
+(2021) also found that the Fluorosense performed well but overestimated chlorophyll *a* at higher algal concentrations.
 Therefore, using this particular instrument to enforce set criteria across a variety of waterbodies would probably not be recommended without further research.
 
-The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations compared to extracted concentrations.
-However, the output from these instruments is not a direct measure of phycocyanin.
+The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations compared to extracted concentrations; however, the output from these instruments is not a direct measure of phycocyanin.
 Instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria, which makes a comparision with other instruments that directly measure phycocyanin difficult.
@@
 A study by Thomson-Laing et al.
-did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (@thomson-laing_predicting_2020).
+(2020) did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures.
 They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.
@@
 Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran.
-Phycocyanin concentrations and cell counts of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012).
+Phycocyanin concentrations and cell counts of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were causing underestimation of the in vivo concentrations of chlorophyll (@zamyadi_monitoring_2012).
 Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments.
 A study by Silva et al.
-found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (@silva_comparison_2016).
+(2016) found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L.
 
-This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies.
-Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments.
+This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study of cyanoHABS.
+In lieu of extractive techniques, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found in vivo phycocyanin to be better correlated with extracted phycocyanin than correlations between in vivo and extracted chlorophyll a across a variety of instruments.
 Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts.
@@
 
-We are grateful to Hilary Snook, Darryl Keith, Steve Rego, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper.
+We are grateful to Hilary Snook, Darryl Keith, Steve Rego, Kara Godineaux, Marty Chintala, and Tim Gleason for constructive reviews of this paper.
 The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.

### 2024-08-02 manuscript/manuscript.qmd
updated manuscript based on MC comments

@@
 Blue lines represent the best fit regression line.
-R^2^ on plot is the coeffecient of determination for this line.](images/fig2_chla_scatter.jpg){#fig-chlascatter width="8.5in" height="5.25in"}
+R^2^ on plot is the coefficient of determination for this line.](images/fig2_chla_scatter.jpg){#fig-chlascatter width="8.5in" height="5.25in"}
 
@@
 #| label: tbl-chlsummary
-#| tbl-cap: "Summary of extracted chlorophyll from Turner Trilogy by waterbody."
+#| tbl-cap: "Summary (mean and standard deviation) of extracted chlorophyll from Turner Trilogy for waterbodies sampled in Rhode Island."
 
@@
 
-![Comparison of fluorometer measurements of phycocyanin by waterbody.
+![Comparison of fluorometer measurements of phycocyanin by Rhode Island waterbody.
 Blue lines represent the best fit regression line.
-R^2^ on plot is the coeffecient of determination for this line.](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width="8.5in" height="5.25in"}
+R^2^ on plot is the coefficient of determination for this line.](images/fig3_phyco_scatter.jpg){#fig-phycoscatter width="8.5in" height="5.25in"}
 
@@
 #| label: tbl-phycosummary
-#| tbl-cap: "Summary of extracted phycocyanin from Turner Trilogy by waterbody."
+#| tbl-cap: "Summary (mean and standard deviation) of extracted phycocyanin from Turner Trilogy for waterbodies sampled in Rhode Island."
 
@@
 
-![Compare fluorometer chlorophyll to cyanobacteria cell counts.
+![Comparison of fluorometer chlorophyll to cyanobacteria cell counts.
 Blue lines represent the best fit regression line.
-R^2^ on plot is the coeffecient of determination for this line.](images/fig5_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
+R^2^ on plot is the coefficient of determination for this line.](images/fig5_chl_vs_cells.jpg){#fig-chlcount width="8.5in" height="5.25in"}
 
-![Compare fluorometer phycocyanin to cyanobacteria cell counts.
+![Comparison of fluorometer phycocyanin to cyanobacteria cell counts.
 Blue lines represent the best fit regression line.
-R^2^ on plot is the coeffecient of determination for this line.](images/fig6_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
+R^2^ on plot is the coefficient of determination for this line.](images/fig6_phyco_vs_cells.jpg){#fig-phycocount width="8.5in" height="5.25in"}
 
